---
title: "__NOMO_REPORT_TITLE__"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    self_contained: true
    mathjax: null
params:
  run: null
  report_title: "nomologR reproducible analysis report"
  include_plots: true
  include_session: true
  max_table_rows: 50
  generated_at: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  results = "asis",
  comment = NA,
  fig.width = 8,
  fig.height = 5.2,
  dpi = 110
)

run <- params$run

html_escape <- function(x) {
  x <- as.character(x)
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;", x, fixed = TRUE)
  x <- gsub(">", "&gt;", x, fixed = TRUE)
  x <- gsub('"', "&quot;", x, fixed = TRUE)
  x
}

emit_note <- function(text) {
  cat(
    '<div class="nomo-note">',
    html_escape(text),
    "</div>\n",
    sep = ""
  )
}

emit_summary <- function(obj) {
  txt <- nomologR:::nomo_report_component_summary_text(obj)
  cat(
    '<details class="nomo-summary-details">',
    '<summary>Detailed component summary</summary>',
    '<pre class="nomo-summary">',
    html_escape(txt),
    "</pre>",
    "</details>\n",
    sep = ""
  )
}

emit_table <- function(x,
                       max_rows = params$max_table_rows,
                       empty = "No rows are available for this section.",
                       truncate = TRUE) {
  x <- nomologR:::nomo_report_flatten_table(x)

  if (!nrow(x)) {
    emit_note(empty)
    return(invisible(NULL))
  }

  n_total <- nrow(x)
  if (isTRUE(truncate) && is.finite(max_rows) && n_total > max_rows) {
    x <- utils::head(x, max_rows)
  }

  html <- knitr::kable(
    x,
    format = "html",
    digits = 3,
    escape = TRUE,
    row.names = FALSE,
    table.attr = 'class="table table-striped table-condensed nomo-table"'
  )
  cat('<div class="table-responsive">\n')
  cat(paste(html, collapse = "\n"), "\n")
  cat("</div>\n")

  if (isTRUE(truncate) && n_total > nrow(x)) {
    emit_note(sprintf(
      "Showing %d of %d rows. The underlying nomo_run object retains all rows.",
      nrow(x),
      n_total
    ))
  }

  invisible(NULL)
}

emit_code <- function(text) {
  if (is.null(text) || !length(text) || !nzchar(as.character(text)[1L])) return()
  cat(
    "<pre><code>",
    html_escape(paste(as.character(text), collapse = "\n")),
    "</code></pre>\n",
    sep = ""
  )
}

emit_plot <- function(obj, type = NULL) {
  if (!isTRUE(params$include_plots)) return(invisible(NULL))

  ans <- nomologR:::nomo_report_safe_plot(obj, type)
  if (isTRUE(ans$ok)) {
    print(ans$plot)
  } else {
    emit_note(paste("Plot unavailable:", ans$message))
  }

  invisible(NULL)
}

stage_status <- nomologR:::nomo_report_run_table(run, "stages")
workflow_decisions <- nomologR:::nomo_report_run_table(run, "decisions")
outstanding <- nomologR:::nomo_report_run_table(run, "requests")
recipe <- nomologR:::nomo_report_run_table(run, "recipe")
settings <- nomologR:::nomo_report_run_table(run, "settings")
scales <- nomologR:::nomo_report_run_table(run, "scales")
trace <- nomologR:::nomo_report_evidence_trace(run)
flagged_trace <- nomologR:::nomo_report_flagged_trace(run)
deviations <- nomologR:::nomo_report_deviations(run)
```

<style>
body { max-width: 1250px; }
.nomo-banner {
  border-left: 5px solid #555;
  padding: 0.8em 1em;
  margin: 1em 0 1.5em 0;
  background: #f6f6f6;
}
.nomo-note {
  border-left: 4px solid #999;
  background: #fafafa;
  padding: 0.6em 0.9em;
  margin: 0.7em 0;
}
.nomo-summary {
  white-space: pre-wrap;
  background: #f8f8f8;
  border: 1px solid #ddd;
  padding: 0.8em;
  margin-top: 0.7em;
}
.nomo-summary-details {
  margin: 0.8em 0 1.1em 0;
}
.nomo-summary-details summary {
  cursor: pointer;
  font-weight: 600;
}
.nomo-table {
  width: 100%;
  font-size: 0.92em;
}
.nomo-table th,
.nomo-table td {
  vertical-align: top !important;
  overflow-wrap: anywhere;
}
.table-responsive {
  width: 100%;
  overflow-x: auto;
  margin-bottom: 0.8em;
}
details {
  margin: 1em 0;
}
</style>

**Generated:** `r params$generated_at`  
**nomologR version:** `r as.character(utils::packageVersion("nomologR"))`  
**Workflow status:** `r run$status`  
**Sample design:** `r run$sample_design`

<div class="nomo-banner">
<strong>Interpretation contract.</strong>
This report archives evidence and researcher decisions. It does not turn review
references into universal pass/fail rules, does not silently delete items or
cases, does not free parameters, and does not respecify models.
</div>

# Researcher inputs and data characteristics

## Workflow state

```{r}
emit_table(stage_status, empty = "No stage-status table is available.")
```

## Scale definitions

```{r}
emit_table(scales, empty = "No scale definitions are available.")
```

## Sample/data roles

```{r}
emit_table(
  nomologR:::nomo_report_data_characteristics(run),
  empty = "Data characteristics could not be reconstructed."
)
```

`r if (identical(run$sample_design, "same_sample")) "Exploratory and confirmatory stages used the same sample. Confirmatory evidence should therefore not be described as independent holdout confirmation." else "Calibration and validation roles were retained explicitly by the workflow."`

## Configured stage settings

```{r}
emit_table(settings, empty = "No stage-specific settings were configured.")
```

## Component recipe

```{r}
emit_table(recipe, empty = "No component recipe is available.")
```

# Item audit

```{r item-audit, results='asis'}
screen <- nomologR:::nomo_report_component_list(run, "screen")

if (!length(screen)) {
  emit_note("The screening stage was not completed in this workflow.")
} else {
  for (scope in names(screen)) {
    obj <- screen[[scope]]
    cat("\n## Scale: ", html_escape(scope), "\n\n", sep = "")
    emit_summary(obj)

    cat("\n### Item summary\n\n")
    emit_table(
      nomologR:::nomo_report_component_table(obj, "screen", "primary")
    )

    cat("\n### Relationship summary\n\n")
    emit_table(
      nomologR:::nomo_report_component_table(obj, "screen", "relationships"),
      empty = "No relationship diagnostics are available."
    )

    log <- nomologR:::nomo_report_component_table(obj, "screen", "decision_log")
    if (nrow(log) && "severity" %in% names(log)) {
      log <- log[tolower(log$severity) %in% c("review", "concern"), , drop = FALSE]
    }
    cat("\n### Review/concern evidence\n\n")
    emit_table(log, empty = "No screening review/concern rows were recorded.")

    emit_plot(obj, "evidence")
  }
}
```

# Factor-retention evidence

```{r factor-retention, results='asis'}
factors <- nomologR:::nomo_report_component_list(run, "factors")

if (!length(factors)) {
  emit_note("Factor-retention evidence was not completed in this workflow.")
} else {
  for (scope in names(factors)) {
    obj <- factors[[scope]]
    cat("\n## Scale: ", html_escape(scope), "\n\n", sep = "")
    emit_summary(obj)

    cat("\n### Retention criteria\n\n")
    emit_table(
      nomologR:::nomo_report_component_table(obj, "factors", "primary")
    )

    cat("\n### Criterion availability and qualifications\n\n")
    emit_table(
      nomologR:::nomo_report_component_table(obj, "factors", "criteria")
    )

    cat("\n### Supporting adequacy evidence\n\n")
    emit_table(
      nomologR:::nomo_report_component_table(obj, "factors", "adequacy")
    )

    emit_plot(obj, "evidence")
  }
}
```

# Exploratory factor analysis

```{r efa, results='asis'}
efa <- nomologR:::nomo_report_component_list(run, "efa")

if (!length(efa)) {
  emit_note("EFA was not completed in this workflow.")
} else {
  for (scope in names(efa)) {
    obj <- efa[[scope]]
    cat("\n## Scale: ", html_escape(scope), "\n\n", sep = "")
    emit_summary(obj)

    cat("\n### Item-level structure\n\n")
    emit_table(
      nomologR:::nomo_report_component_table(obj, "efa", "primary")
    )

    cat("\n### Pattern matrix\n\n")
    emit_table(
      nomologR:::nomo_report_component_table(obj, "efa", "pattern")
    )

    cat("\n### Largest localized residuals\n\n")
    residuals <- nomologR:::nomo_report_component_table(obj, "efa", "residuals")
    if (nrow(residuals)) residuals <- utils::head(residuals, 10L)
    emit_table(residuals, empty = "No localized EFA residual table is available.")

    emit_plot(obj, "items")
  }
}
```

# Confirmatory factor analysis

```{r cfa, results='asis'}
cfa_list <- nomologR:::nomo_report_component_list(run, "cfa")

if (!length(cfa_list)) {
  emit_note("CFA was not completed in this workflow.")
} else {
  cfa <- cfa_list[[1L]]
  emit_summary(cfa)

  if (!is.null(run$decisions$cfa_model$value)) {
    cat("\n## Researcher-specified CFA syntax\n\n")
    emit_code(run$decisions$cfa_model$value)
  }

  cat("\n## Global fit evidence\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(cfa, "cfa", "primary")
  )

  cat("\n## Standardized loadings\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(cfa, "cfa", "loadings")
  )

  cat("\n## Improper-solution / Heywood evidence\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(cfa, "cfa", "heywood"),
    empty = "No configured improper-solution/Heywood rows were recorded."
  )

  cat("\n## Modification indices - diagnostic only\n\n")
  mi <- nomologR:::nomo_report_component_table(
    cfa,
    "cfa",
    "modification_indices"
  )
  if (nrow(mi)) mi <- utils::head(mi, 10L)
  emit_table(
    mi,
    empty = "No modification-index table is available."
  )
  emit_note(
    "Modification indices are post-hoc diagnostic evidence. They do not authorize automatic model respecification."
  )

  emit_plot(cfa, "fit")
  emit_plot(cfa, "loadings")
}
```

# Reliability

```{r reliability, results='asis'}
rel_list <- nomologR:::nomo_report_component_list(run, "reliability")

if (!length(rel_list)) {
  emit_note("Reliability evidence was not completed in this workflow.")
} else {
  rel <- rel_list[[1L]]
  emit_summary(rel)

  cat("\n## Reliability coefficients\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(rel, "reliability", "primary")
  )

  cat("\n## Alpha availability / estimand context\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(rel, "reliability", "alpha_status"),
    empty = "No secondary-alpha status table is available."
  )

  emit_plot(rel)
}
```

# Convergent and discriminant evidence

```{r validity, results='asis'}
val_list <- nomologR:::nomo_report_component_list(run, "validity")

if (!length(val_list)) {
  emit_note("Convergent/discriminant evidence was not completed in this workflow.")
} else {
  val <- val_list[[1L]]
  emit_summary(val)

  cat("\n## Convergent evidence\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(val, "validity", "primary")
  )

  cat("\n## Construct-separation evidence\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(val, "validity", "discriminant"),
    empty = "No pairwise construct-separation table is available."
  )

  cat("\n## HTMT-family availability\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(val, "validity", "htmt_status"),
    empty = "No HTMT-family status table is available."
  )

  emit_plot(val, "ave")
  if (nrow(nomologR:::nomo_report_component_table(
    val, "validity", "discriminant"
  ))) {
    emit_plot(val, "discriminant")
  }
}
```

# Measurement invariance

```{r invariance, results='asis'}
inv_list <- nomologR:::nomo_report_component_list(run, "invariance")

if (!length(inv_list)) {
  status <- stage_status$status[stage_status$stage == "invariance"]
  if (length(status) && identical(status[[1L]], "not_requested")) {
    emit_note("Measurement invariance was not requested in this workflow.")
  } else {
    emit_note("Measurement invariance was not completed in this workflow.")
  }
} else {
  inv <- inv_list[[1L]]
  emit_summary(inv)

  cat("\n## Fit across invariance levels\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(inv, "invariance", "primary")
  )

  cat("\n## Researcher-specified partial releases\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(inv, "invariance", "partial"),
    empty = "No partial-invariance releases were supplied."
  )

  cat("\n## Local equality-constraint strain\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(inv, "invariance", "local_strain"),
    empty = "No localized equality-constraint diagnostics are available."
  )

  emit_plot(inv, "change")
}
```

# Nomological network

```{r network, results='asis'}
net_list <- nomologR:::nomo_report_component_list(run, "network")

if (!length(net_list)) {
  status <- stage_status$status[stage_status$stage == "network"]
  if (length(status) && identical(status[[1L]], "not_requested")) {
    emit_note("A theory-specified nomological network was not requested.")
  } else {
    emit_note("The nomological-network stage was not completed.")
  }
} else {
  net <- net_list[[1L]]
  emit_summary(net)

  cat("\n## Theory-specified relation evidence\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(net, "network", "primary")
  )

  cat("\n## Structural relation/model provenance\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(net, "network", "relations")
  )

  cat("\n## Replication evidence\n\n")
  replication <- nomologR:::nomo_report_component_table(
    net, "network", "replication"
  )
  emit_table(
    replication,
    empty = "No validation-sample replication evidence is available."
  )

  emit_plot(net, "effects")
  if (nrow(replication)) emit_plot(net, "replication")
}
```

# Decision log

## Workflow decisions

```{r}
emit_table(
  workflow_decisions,
  empty = "No explicit workflow decisions were recorded."
)
```

## Outstanding decision requests

```{r}
emit_table(
  outstanding,
  empty = "No consequential researcher decisions remain outstanding."
)
```

## Review and concern evidence

Each row below carries a stable report-local **evidence ID**. The metric,
reference, observation, and recommendation remain together so autogenerated
guidance can be traced back to the evidence that produced it.

```{r}
emit_table(
  flagged_trace,
  empty = "No component review/concern evidence rows were recorded."
)
```

# Deviations and post-hoc decisions

```{r}
emit_table(
  deviations,
  empty = paste(
    "No researcher-specified partial-invariance releases, post-hoc network",
    "relations, or explicit workflow revision decisions were identified."
  )
)
```

A diagnostic suggestion (for example, a CFA modification index) is not labeled
as a deviation unless the researcher actually makes and records a model change.

# Methods and citations

The report reflects `nomologR` orchestration plus established statistical
engines. `installed_version` records the package version used in this rendering
session; package-provided citation text may identify a different citation
release. Cite the software/method packages appropriate to the analyses actually
reported in a manuscript.

```{r}
emit_table(nomologR:::nomo_report_citations(), max_rows = Inf)
```

# Reproducibility

## Workflow call history

```{r}
emit_table(
  nomologR:::nomo_report_call_history(run),
  max_rows = Inf,
  empty = "No call history is available."
)
```

## Package versions

```{r}
emit_table(nomologR:::nomo_report_package_versions(), max_rows = Inf)
```

## Full evidence trace

This appendix is intentionally **not truncated**. Each autogenerated
recommendation remains on the same row as its evidence ID, component, metric,
reference, severity, and observation.

<details>
<summary>Show full evidence trace</summary>

```{r}
emit_table(
  trace,
  max_rows = Inf,
  empty = "No component evidence log rows are available.",
  truncate = FALSE
)
```

</details>

## Session information

```{r session-info, results='asis'}
if (isTRUE(params$include_session)) {
  cat(
    '<pre class="nomo-summary">',
    html_escape(nomologR:::nomo_report_session_text()),
    "</pre>\n",
    sep = ""
  )
} else {
  emit_note("Full session information was disabled for this report.")
}
```

---

**Report boundary:** this document is a reproducible presentation of the
supplied `nomo_run` object. It does not replace substantive theory, content
validity, design justification, or researcher judgment.
