---
title: "__NOMO_REPORT_TITLE__"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    self_contained: true
    mathjax: null
params:
  run: null
  report_title: "nomologR reproducible analysis report"
  include_plots: true
  include_session: true
  apa_tables: false
  max_table_rows: 50
  generated_at: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  results = "asis",
  comment = NA,
  fig.width = 8,
  fig.height = 5.2,
  dpi = 110
)

run <- params$run

# Item and scale names are the researcher's, so table headings keep them as
# written rather than turning them into words (#89).
protected_names <- unique(c(names(run$scales), unlist(run$scales, use.names = FALSE)))

# A Word document cannot carry raw HTML: pandoc drops it when writing .docx,
# which would lose every table and the interpretation contract. Each emitter
# below writes HTML for an HTML report and markdown for a Word one, so the HTML
# report is unchanged.
word_output <- isTRUE(knitr::pandoc_to("docx"))

# Pandoc reads markdown wherever text is written, even inside an HTML table, so
# every heading, note, and cell built from data is escaped to show exactly as
# written (#145): lavaan's `~~` would otherwise vanish and `a*x1` lose its
# asterisk.
text_escape <- function(x) {
  nomologR:::nomo_report_escape(x, if (word_output) "markdown" else "html")
}

# For the verbatim blocks, which pandoc does not read as markdown.
html_escape <- function(x) {
  x <- as.character(x)
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;", x, fixed = TRUE)
  x <- gsub(">", "&gt;", x, fixed = TRUE)
  x <- gsub('"', "&quot;", x, fixed = TRUE)
  x
}

# The headings, cells, and notes the report writes, so the closing table can
# define every abbreviation it shows (#144).
shown <- new.env()
shown$labels <- character()
shown$text <- character()
remember <- function(labels = character(), text = character()) {
  shown$labels <- c(shown$labels, labels)
  shown$text <- c(shown$text, text)
  invisible(NULL)
}

emit_heading <- function(text, level = 2L) {
  remember(text = text)
  cat("\n", strrep("#", level), " ", text_escape(text), "\n\n", sep = "")
}

emit_note <- function(text) {
  remember(text = text)
  if (word_output) {
    cat("\n*", text_escape(text), "*\n\n", sep = "")
    return(invisible(NULL))
  }
  cat(
    '<div class="nomo-note">',
    text_escape(text),
    "</div>\n",
    sep = ""
  )
}

emit_summary <- function(obj) {
  txt <- nomologR:::nomo_report_component_summary_text(obj)
  if (word_output) {
    cat("\n```\n", txt, "\n```\n\n", sep = "")
    return(invisible(NULL))
  }
  cat(
    '<details class="nomo-summary-details">',
    '<summary>Detailed component summary</summary>',
    '<pre class="nomo-summary">',
    html_escape(txt),
    "</pre>",
    "</details>\n",
    sep = ""
  )
}

# Values are rounded and labelled by nomo_report_table(), which also escapes
# each cell. `markdown = TRUE` keeps the markdown of trusted text, such as the
# italics in R's citations.
emit_table <- function(x,
                       max_rows = params$max_table_rows,
                       empty = "No rows are available for this section.",
                       truncate = TRUE,
                       markdown = FALSE) {
  table <- nomologR:::nomo_report_table(
    x,
    protected = protected_names,
    word = word_output,
    max_rows = if (isTRUE(truncate)) max_rows else Inf,
    markdown = markdown
  )

  if (is.null(table)) {
    emit_note(empty)
    return(invisible(NULL))
  }
  remember(table$labels, table$cells)

  if (word_output) {
    cat("\n", table$markup, "\n\n", sep = "")
  } else {
    cat('<div class="table-responsive">\n')
    cat(table$markup, "\n")
    cat("</div>\n")
  }

  for (note in table$notes) emit_note(note)

  invisible(NULL)
}

emit_code <- function(text) {
  if (is.null(text) || !length(text) || !nzchar(as.character(text)[1L])) return()
  if (word_output) {
    cat("\n```\n", paste(as.character(text), collapse = "\n"), "\n```\n\n", sep = "")
    return(invisible(NULL))
  }
  cat(
    "<pre><code>",
    html_escape(paste(as.character(text), collapse = "\n")),
    "</code></pre>\n",
    sep = ""
  )
}

emit_plot <- function(obj, type = NULL) {
  if (!isTRUE(params$include_plots)) return(invisible(NULL))

  ans <- nomologR:::nomo_report_safe_plot(obj, type)
  if (isTRUE(ans$ok)) {
    print(ans$plot)
    # knitr writes the image without a trailing blank line, so a heading
    # emitted next would be read as part of the image's paragraph and appear
    # as literal "## ..." text (#89).
    cat("\n\n")
  } else {
    emit_note(paste("Plot unavailable:", ans$message))
  }

  invisible(NULL)
}

stage_status <- nomologR:::nomo_report_run_table(run, "stages")
workflow_decisions <- nomologR:::nomo_report_run_table(run, "decisions")
outstanding <- nomologR:::nomo_report_run_table(run, "requests")
recipe <- nomologR:::nomo_report_run_table(run, "recipe")
settings <- nomologR:::nomo_report_run_table(run, "settings")
scales <- nomologR:::nomo_report_run_table(run, "scales")
trace <- nomologR:::nomo_report_evidence_trace(run)
flagged_trace <- nomologR:::nomo_report_flagged_trace(run)
lineage <- nomologR:::nomo_report_lineage(run)
deviations <- nomologR:::nomo_report_deviations(run)
```

<style>
body { max-width: 1250px; }
.nomo-banner {
  border-left: 5px solid #555;
  padding: 0.8em 1em;
  margin: 1em 0 1.5em 0;
  background: #f6f6f6;
}
.nomo-note {
  border-left: 4px solid #999;
  background: #fafafa;
  padding: 0.6em 0.9em;
  margin: 0.7em 0;
}
.nomo-summary {
  white-space: pre-wrap;
  background: #f8f8f8;
  border: 1px solid #ddd;
  padding: 0.8em;
  margin-top: 0.7em;
}
.nomo-summary-details {
  margin: 0.8em 0 1.1em 0;
}
.nomo-summary-details summary {
  cursor: pointer;
  font-weight: 600;
}
.nomo-table {
  width: 100%;
  font-size: 0.92em;
}
.nomo-table th,
.nomo-table td {
  vertical-align: top !important;
  overflow-wrap: anywhere;
}
.table-responsive {
  width: 100%;
  overflow-x: auto;
  margin-bottom: 0.8em;
}
details {
  margin: 1em 0;
}
/* APA 7 tables (#145): a rule above the table, one beneath the column
   headings, and one below the table, and none between body rows. The page
   script gives every table Bootstrap's classes, whose row borders these
   replace. */
.nomo-apa table {
  width: auto;
  border-top: 2px solid #000;
  border-bottom: 2px solid #000;
  margin-bottom: 0.5em;
}
.nomo-apa table > thead > tr > th {
  border-top: none !important;
  border-bottom: 1px solid #000 !important;
}
.nomo-apa table > tbody > tr > td {
  border-top: none !important;
}
</style>

**Generated:** `r params$generated_at`  
**nomologR version:** `r as.character(utils::packageVersion("nomologR"))`  
**Workflow status:** `r nomologR:::nomo_present_status(run$status)`  
**Sample design:** `r gsub("_", " ", run$sample_design, fixed = TRUE)`

```{r interpretation-contract}
contract <- paste(
  "This report archives evidence and researcher decisions. It does not turn review",
  "references into universal pass/fail rules, does not silently delete items or",
  "cases, does not free parameters, and does not respecify models."
)
if (word_output) {
  cat("> **Interpretation contract.** ", contract, "\n\n", sep = "")
} else {
  cat(
    '<div class="nomo-banner">\n<strong>Interpretation contract.</strong>\n',
    contract, "\n</div>\n",
    sep = ""
  )
}
emit_note(nomologR:::nomo_report_reading_note())
```

# Researcher inputs and data characteristics

## Workflow state

```{r}
emit_table(stage_status, empty = "No stage-status table is available.")
```

## Scale definitions

```{r}
emit_table(scales, empty = "No scale definitions are available.")
```

```{r content-review, results='asis'}
content_review <- nomologR:::nomo_report_content_review(run)
if (!is.null(content_review)) {
  cat("\n## Content review\n\n")
  emit_note(content_review$summary)
  emit_table(content_review$items, truncate = FALSE)
}
```

## Sample/data roles

```{r}
emit_table(
  nomologR:::nomo_report_data_characteristics(run),
  empty = "Data characteristics could not be reconstructed."
)
```

`r if (identical(run$sample_design, "same_sample")) "Exploratory and confirmatory stages used the same sample. Confirmatory evidence should therefore not be described as independent holdout confirmation." else "Calibration and validation roles were retained explicitly by the workflow."`

## Configured stage settings

```{r}
emit_table(settings, empty = "No stage-specific settings were configured.")
```

## Component recipe

```{r}
emit_table(recipe, empty = "No component recipe is available.")
```

# Item audit

```{r item-audit, results='asis'}
screen <- nomologR:::nomo_report_component_list(run, "screen")

if (!length(screen)) {
  emit_note("The screening stage was not completed in this workflow.")
} else {
  for (scope in names(screen)) {
    obj <- screen[[scope]]
    cat("\n## Scale: ", text_escape(scope), "\n\n", sep = "")
    emit_summary(obj)

    cat("\n### Item summary\n\n")
    emit_table(
      nomologR:::nomo_report_component_table(obj, "screen", "primary")
    )

    cat("\n### Relationship summary\n\n")
    emit_table(
      nomologR:::nomo_report_component_table(obj, "screen", "relationships"),
      empty = "No relationship diagnostics are available."
    )

    log <- nomologR:::nomo_report_flagged_rows(
      nomologR:::nomo_report_component_table(obj, "screen", "decision_log")
    )
    cat("\n### Review/concern evidence\n\n")
    emit_table(log, empty = "No screening review/concern rows were recorded.")

    emit_plot(obj, "evidence")
  }

  if (!is.null(run$results$effort)) {
    effort_report <- nomologR:::nomo_report_effort(run$results$effort)
    cat("\n## Careless responding\n\n")
    emit_note(effort_report$summary)
    emit_table(effort_report$indices, truncate = FALSE)
    cat("\n### Careless-responding evidence\n\n")
    emit_table(effort_report$log, empty = "No careless-responding rows were recorded.")
  }
}
```

# Factor-retention evidence

```{r factor-retention, results='asis'}
factors <- nomologR:::nomo_report_component_list(run, "factors")

if (!length(factors)) {
  emit_note("Factor-retention evidence was not completed in this workflow.")
} else {
  for (scope in names(factors)) {
    obj <- factors[[scope]]
    cat("\n## Scale: ", text_escape(scope), "\n\n", sep = "")
    emit_summary(obj)

    cat("\n### Retention criteria\n\n")
    emit_table(
      nomologR:::nomo_report_component_table(obj, "factors", "primary")
    )

    cat("\n### Criterion availability and qualifications\n\n")
    emit_table(
      nomologR:::nomo_report_component_table(obj, "factors", "criteria")
    )

    cat("\n### Supporting adequacy evidence\n\n")
    emit_table(
      nomologR:::nomo_report_component_table(obj, "factors", "adequacy")
    )

    emit_plot(obj, "evidence")
  }
}
```

# Exploratory factor analysis

```{r efa, results='asis'}
efa <- nomologR:::nomo_report_component_list(run, "efa")

if (!length(efa)) {
  emit_note("EFA was not completed in this workflow.")
} else {
  for (scope in names(efa)) {
    obj <- efa[[scope]]
    cat("\n## Scale: ", text_escape(scope), "\n\n", sep = "")
    emit_summary(obj)

    cat("\n### Item-level structure\n\n")
    emit_table(
      nomologR:::nomo_report_component_table(obj, "efa", "primary")
    )

    cat("\n### Pattern matrix\n\n")
    emit_table(
      nomologR:::nomo_report_component_table(obj, "efa", "pattern")
    )

    cat("\n### Largest localized residuals\n\n")
    residuals <- nomologR:::nomo_report_component_table(obj, "efa", "residuals")
    if (nrow(residuals)) residuals <- utils::head(residuals, 10L)
    emit_table(residuals, empty = "No localized EFA residual table is available.")

    emit_plot(obj, "items")
  }
}
```

# Confirmatory factor analysis

```{r cfa, results='asis'}
cfa_list <- nomologR:::nomo_report_component_list(run, "cfa")

if (!length(cfa_list)) {
  emit_note("CFA was not completed in this workflow.")
} else {
  cfa <- cfa_list[[1L]]
  emit_summary(cfa)

  if (!is.null(run$decisions$cfa_model$value)) {
    emit_heading("Researcher-specified CFA syntax")
    emit_code(run$decisions$cfa_model$value)
  }

  cat("\n## Global fit evidence\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(cfa, "cfa", "primary")
  )

  cat("\n## Standardized loadings\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(cfa, "cfa", "loadings")
  )

  cat("\n## Improper-solution / Heywood evidence\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(cfa, "cfa", "heywood"),
    empty = "No configured improper-solution/Heywood rows were recorded."
  )

  cat("\n## Modification indices - diagnostic only\n\n")
  mi <- nomologR:::nomo_report_component_table(
    cfa,
    "cfa",
    "modification_indices"
  )
  if (nrow(mi)) mi <- utils::head(mi, 10L)
  emit_table(
    mi,
    empty = "No modification-index table is available."
  )
  emit_note(
    "Modification indices are post hoc diagnostic evidence. They do not authorize automatic model respecification."
  )

  emit_plot(cfa, "fit")
  emit_plot(cfa, "loadings")
}
```

# Reliability

```{r reliability, results='asis'}
rel_list <- nomologR:::nomo_report_component_list(run, "reliability")

if (!length(rel_list)) {
  emit_note("Reliability evidence was not completed in this workflow.")
} else {
  rel <- rel_list[[1L]]
  emit_summary(rel)

  cat("\n## Reliability coefficients\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(rel, "reliability", "primary")
  )

  cat("\n## Alpha availability / estimand context\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(rel, "reliability", "alpha_status"),
    empty = "No secondary-alpha status table is available."
  )

  emit_plot(rel)
}
```

# Convergent and discriminant evidence

```{r validity, results='asis'}
val_list <- nomologR:::nomo_report_component_list(run, "validity")

if (!length(val_list)) {
  emit_note("Convergent/discriminant evidence was not completed in this workflow.")
} else {
  val <- val_list[[1L]]
  emit_summary(val)

  cat("\n## Convergent evidence\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(val, "validity", "primary")
  )

  cat("\n## Construct-separation evidence\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(val, "validity", "discriminant"),
    empty = "No pairwise construct-separation table is available."
  )

  emit_heading("HTMT-family availability")
  emit_table(
    nomologR:::nomo_report_component_table(val, "validity", "htmt_status"),
    empty = "No HTMT-family status table is available."
  )

  emit_plot(val, "ave")
  if (nrow(nomologR:::nomo_report_component_table(
    val, "validity", "discriminant"
  ))) {
    emit_plot(val, "discriminant")
  }
}
```

```{r scores-section, results='asis'}
if (!is.null(run$results$scores)) {
  scores_report <- nomologR:::nomo_report_scores(run$results$scores)
  cat("\n# Scores\n\n")
  emit_note(scores_report$summary)
  cat("\n## Score properties (Grice, 2001)\n\n")
  emit_table(scores_report$diagnostics, truncate = FALSE)
  cat("\n## Notes\n\n")
  emit_table(scores_report$notes, empty = "No notes were recorded.", truncate = FALSE)
}
```

# Measurement invariance

```{r invariance, results='asis'}
inv_list <- nomologR:::nomo_report_component_list(run, "invariance")

if (!length(inv_list)) {
  status <- stage_status$status[stage_status$stage == "invariance"]
  if (length(status) && identical(status[[1L]], "not_requested")) {
    emit_note("Measurement invariance was not requested in this workflow.")
  } else {
    emit_note("Measurement invariance was not completed in this workflow.")
  }
} else {
  inv <- inv_list[[1L]]
  emit_summary(inv)

  cat("\n## Fit across invariance levels\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(inv, "invariance", "primary")
  )

  cat("\n## Researcher-specified partial releases\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(inv, "invariance", "partial"),
    empty = "No partial-invariance releases were supplied."
  )

  cat("\n## Local equality-constraint strain\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(inv, "invariance", "local_strain"),
    empty = "No localized equality-constraint diagnostics are available."
  )

  emit_plot(inv, "change")
}
```

# Nomological network

```{r network, results='asis'}
net_list <- nomologR:::nomo_report_component_list(run, "network")

if (!length(net_list)) {
  status <- stage_status$status[stage_status$stage == "network"]
  if (length(status) && identical(status[[1L]], "not_requested")) {
    emit_note("A theory-specified nomological network was not requested.")
  } else {
    emit_note("The nomological-network stage was not completed.")
  }
} else {
  net <- net_list[[1L]]
  emit_summary(net)

  cat("\n## Theory-specified relation evidence\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(net, "network", "primary")
  )

  cat("\n## Structural relation/model provenance\n\n")
  emit_table(
    nomologR:::nomo_report_component_table(net, "network", "relations")
  )

  cat("\n## Replication evidence\n\n")
  replication <- nomologR:::nomo_report_component_table(
    net, "network", "replication"
  )
  emit_table(
    replication,
    empty = "No validation-sample replication evidence is available."
  )

  emit_plot(net, "effects")
  if (nrow(replication)) emit_plot(net, "replication")
}
```

```{r missing-section, results='asis'}
if (length(run$results$missing)) {
  cat("\n# Missing-data sensitivity\n\n")
  targets <- c(cfa = "Measurement model", network = "Nomological network")
  for (target in intersect(names(targets), names(run$results$missing))) {
    missing_report <- nomologR:::nomo_report_missing(run$results$missing[[target]])
    cat("\n## ", targets[[target]], "\n\n", sep = "")
    emit_note(missing_report$summary)
    cat("\n### Strategies\n\n")
    emit_table(missing_report$strategies, truncate = FALSE)
    cat("\n### Largest differences from the reference\n\n")
    emit_table(missing_report$differences, empty = missing_report$differences_empty)
    cat("\n### Fit by strategy\n\n")
    emit_table(missing_report$fit, truncate = FALSE)
    if (!is.null(missing_report$reliability) && nrow(missing_report$reliability)) {
      cat("\n### Reliability by strategy\n\n")
      emit_table(missing_report$reliability, truncate = FALSE)
    }
    cat("\n### Missing-data evidence\n\n")
    emit_table(missing_report$log, truncate = FALSE)
  }
}
```

# Decision log

## Workflow decisions

```{r}
emit_table(
  workflow_decisions,
  empty = "No explicit workflow decisions were recorded."
)
```

## Outstanding decision requests

```{r}
emit_table(
  outstanding,
  empty = "No consequential researcher decisions remain outstanding."
)
```

## Review and concern evidence

Each row below carries a stable report-local **evidence ID**. The metric,
reference, observation, and recommendation remain together so autogenerated
guidance can be traced back to the evidence that produced it.

```{r}
emit_table(
  flagged_trace,
  empty = "No component review/concern evidence rows were recorded."
)
```

# Revision lineage

```{r}
emit_table(
  lineage,
  empty = "This workflow was not produced by revising an earlier workflow."
)
```

Each row records one revision: what changed, whether the change was
prespecified or post hoc, the researcher's rationale, and the comparison
between the parent and revised measurement models. A revision prompted by the
results is evaluated on the data that prompted it, so independent confirmation
is recommended.

# Deviations and post hoc decisions

```{r}
emit_table(
  deviations,
  empty = paste(
    "No researcher-specified partial-invariance releases, post hoc network",
    "relations, or explicit workflow revision decisions were identified."
  )
)
```

A diagnostic suggestion (for example, a CFA modification index) is not labeled
as a deviation unless the researcher actually makes and records a model change.

# Methods and citations

## Methods used in this workflow

Every method below was used in producing this run. Methods the package can
perform but this workflow did not use are deliberately absent, so this table
describes what happened rather than what the package is capable of.

`lineage` places a method in the literature: `contemporary` is current
recommended practice, `historical` is a technique retained so published work
can be recognized and interpreted, and `emerging` is a recent proposal with a
smaller evidence base. `role` says how the evidence was used: `primary`,
`supporting`, or `context`. A `context` method was displayed for recognition
only and took no part in any synthesis or decision.

```{r}
emit_table(
  nomologR:::nomo_report_methods(run),
  max_rows = Inf,
  empty = "No registry methods were recorded for this workflow."
)
```

## Method references

```{r}
emit_table(
  nomologR:::nomo_report_method_references(run),
  max_rows = Inf,
  empty = "No method references are available.",
  markdown = TRUE
)
```

## Software

The report reflects `nomologR` orchestration plus established statistical
engines. `installed_version` records the package version used in this rendering
session; package-provided citation text may identify a different citation
release.

```{r}
emit_table(nomologR:::nomo_report_citations(), max_rows = Inf, markdown = TRUE)
```

```{r manuscript-tables, results='asis'}
if (isTRUE(params$apa_tables)) {
  apa <- nomologR:::nomo_report_apa_tables(run)
  cat("\n# Manuscript tables\n\n")
  if (!length(apa)) {
    emit_note("This workflow holds no results with an APA table yet.")
  } else {
    emit_note(paste(
      "APA 7 tables for the results in this report, ready to copy into a",
      "manuscript. They follow the reference-value language of the report:",
      "no cell reads pass or fail."
    ))
    # In HTML the tables carry APA rules (top, beneath the headings, bottom)
    # rather than the page's rule under every row.
    if (!word_output) cat("\n::: {.nomo-apa}\n\n")
    for (tab in apa) {
      cat("\n", paste(nomologR:::nomo_apa_markdown(tab), collapse = "\n"), "\n\n", sep = "")
    }
    if (!word_output) cat("\n:::\n\n")
  }
}
```

# Reproducibility

## Workflow call history

```{r}
# As R code: in a table its quotes became typographic ones (#145).
call_code <- nomologR:::nomo_report_call_code(run)
if (nzchar(call_code)) {
  emit_code(call_code)
} else {
  emit_note("No call history is available.")
}
```

## Package versions

```{r}
emit_table(nomologR:::nomo_report_package_versions(), max_rows = Inf)
```

## Full evidence trace

This appendix is intentionally **not truncated**. Each autogenerated
recommendation remains on the same row as its evidence ID, component, metric,
reference, severity, and observation.

```{r evidence-trace-open}
if (!word_output) cat("<details>\n<summary>Show full evidence trace</summary>\n")
```

```{r}
emit_table(
  trace,
  max_rows = Inf,
  empty = "No component evidence log rows are available.",
  truncate = FALSE
)
```

```{r evidence-trace-close}
if (!word_output) cat("</details>\n")
```

## Session information

```{r session-info, results='asis'}
if (isTRUE(params$include_session) && word_output) {
  emit_code(nomologR:::nomo_report_session_text())
} else if (isTRUE(params$include_session)) {
  cat(
    '<pre class="nomo-summary">',
    html_escape(nomologR:::nomo_report_session_text()),
    "</pre>\n",
    sep = ""
  )
} else {
  emit_note("Full session information was disabled for this report.")
}
```

# Abbreviations

```{r abbreviations}
emit_table(
  nomologR:::nomo_report_abbreviations(shown$labels, shown$text),
  max_rows = Inf,
  empty = "The report's tables use no abbreviation."
)
```

---

**Report boundary:** this document is a reproducible presentation of the
supplied `nomo_run` object. It does not replace substantive theory, content
validity, design justification, or researcher judgment.
