---
title: "From content review to empirical screening"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From content review to empirical screening}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(nomologR)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)

# A corrected item-rest correlation from a screen, for the prose.
item_rest <- function(screen, item) {
  r <- screen$relationship_summary$corrected_item_rest_r
  sprintf("%.2f", r[screen$relationship_summary$item == item])
}
```

Content review and an empirical screen answer different questions. An expert
panel judges whether an item represents the construct as it was defined. Response
data show how the item behaves: whether it varies, what it shares with the other
items, and whether it works the same way in different groups. Neither answer
contains the other, and an item can pass one and fail the other.

This article takes one item set across that boundary. A content review sorted
twelve items and carried ten of them forward. Here those ten are screened on
400 simulated responses. The content-review half, meaning how the panel's sort
was analyzed and how the handoff was made, is the joint walkthrough in
`contentvalidR`,
[One item set, both stages](https://juhalt.github.io/contentvalidR/articles/one-item-set-both-stages.html).
Both articles use the same data. The data are simulated from a model that is
documented in `?nomo_demo_walkthrough`, so every claim below can be checked
against the truth.

## What content review decided

`contentvalidR` records its decisions in a *handoff*. This package stores the
handoff for these items, so the article runs without `contentvalidR` installed:

```{r handoff}
handoff <- readRDS(
  system.file("extdata", "content-handoff-walkthrough.rds", package = "nomologR")
)
handoff$item_evidence[, c("item", "scale", "status", "carried", "keying")]
```

The panel carried ten items. It held back `EF5` and `TF5`, whose sorts did not
meet the criterion. It placed the carried items in two facets:

* Effort Regulation (`EF`), keeping going when the work is dull or hard;
* Task Focus (`TF`), staying on the task rather than switching away.

Keying `-1` marks `EF2` and `TF2` as reverse-worded: a high answer means *less*
persistence.

## The responses, and what each item was built to do

```{r roles}
nomo_demo_walkthrough_items[, c("item", "role")]
```

A real study does not come with a `role` column. Here it is known because the
data were simulated to contain specific problems, and it is the answer key for
what follows.

## Screening the data as they arrive

`nomo_run()` takes the handoff whole: its carried items and facets become the
run's scales, and each facet is screened on its own.

```{r raw-run}
run <- nomo_run(nomo_demo_walkthrough, scales = handoff)
run
```

Every item is flagged. The Effort Regulation audit shows why:

```{r raw-ef}
summary(run$results$screen$EF)
```

`EF2` correlates negatively with the rest of its facet. The audit says what that
most likely means: the item is declared reverse-keyed, and recoded as declared
its correlation would be positive. The negative sign is a coding matter, not
evidence against the item.

Left as answered, the item also makes the rest of the facet look bad. Every other EF
item's correlation with the rest of the scale includes `EF2` running the wrong
way, so all of them fall below the teaching reference. A screen of data that are
not yet recoded cannot say much about any item in the facet.

`nomologR` never recodes data. Recoding is the researcher's step, and it belongs
in the analysis code where it can be seen.

## Recoding as content review declared

```{r recode}
walk <- nomo_demo_walkthrough
reversed <- handoff$item_evidence$item[handoff$item_evidence$keying %in% -1]
walk[reversed] <- 6L - walk[reversed]
```

The responses are now recoded, so the run should not treat any item as still
needing it. `reverse = character(0)` says so, in place of the handoff's keying:

```{r recoded-run}
run <- nomo_run(
  walk,
  scales = handoff,
  settings = list(screen = list(reverse = character(0), scale_range = c(1, 5)))
)
ef <- run$results$screen$EF
summary(ef)
```

Recoded, `EF2` correlates `r item_rest(ef, "EF2")` with the rest of its facet,
and one EF item remains flagged: `EF4`, at `r item_rest(ef, "EF4")`. Eighteen of
twenty judges placed `EF4` in Effort Regulation, and they were right about its
wording. Keeping to a self-set study schedule reads as effort. In the responses,
though, it shares almost nothing with the other items. This is what content
review cannot see.

`EF3` is not flagged here (`r item_rest(ef, "EF3")`), but its distribution stands
out: the "Top share" column shows most answers in a single category. Almost
everyone finishes the assignments that count toward their grade. The next
stage shows what that costs.

## The measurement model

An exploratory factor analysis of the ten recoded items, with the two facets the
panel defined:

```{r efa}
efa <- nomo_efa(walk[handoff$items], factors = 2)
summary(efa)
```

Three items stand apart:

* `EF4` has almost no common variance, as the screen suggested.
* `EF3` loads weakly as well. Its answers are piled at the top of the scale,
  and a variable that barely varies cannot correlate strongly with anything.
* `TF4` loads on both factors. Working through a task without taking breaks is
  effort as much as focus. The panel saw one facet, and the data see two.

A confirmatory model with the panel's structure locates the same strain:

```{r cfa}
cfa <- nomo_cfa(nomo_model(handoff$scales), data = walk)
cfa
head(nomo_table(cfa, "modification_indices"), 3)
```

The largest modification index asks for `TF4` to load on Effort Regulation too.
It is the same item the exploratory analysis flagged, found from the other
direction. As always in this package, the index locates strain; it does not free
the parameter.

## Two cohorts

The respondents came from two cohorts. Before comparing their means, check that
the items measure the same thing in both:

```{r invariance}
inv <- nomo_invariance(
  nomo_model(handoff$scales),
  data = walk,
  group = "cohort",
  levels = c("configural", "metric", "scalar")
)
summary(inv)
```

Equal loadings hold; equal intercepts cost fit. The largest score diagnostic
is `TF6`'s intercept. Putting a phone away while studying is a specific
behavior, and the two cohorts faced different classroom rules about phones. An
item can be content-valid and still not comparable across groups. Whether to
release that intercept (see `nomo_partial()`) is a researcher decision, and it
needs a rationale such as that one.

## Where the two stages disagree

| Item | Content review | Empirical evidence | What it teaches |
|---|---|---|---|
| `EF4` | Carried: 18 of 20 judges | Almost no common variance | Evidence against the item that no panel could see. Revise it, or drop it with a rationale. |
| `EF3` | Carried: 18 of 20 judges | Weak loading, from answers piled at the ceiling | The screen is not the final word. `EF3` is the only item about finishing required work; dropping it narrows the domain the panel defined. |
| `TF4` | Carried: 16 of 20 judges | Loads on both facets | A question about content. Revise the wording, or model the cross-loading with a rationale. |
| `TF6` | Carried: 18 of 20 judges | Intercept differs by cohort | Explain the difference before comparing cohort means. |
| `EF2`, `TF2` | Carried; reverse-worded | Negative until recoded | Coding, not evidence. |
| `EF5`, `TF5` | Held back | Not screened | They remain in `nomo_demo_walkthrough`, so you can see what keeping them would have done. |

Neither stage overrules the other. `EF4` is the case for not trusting a panel
alone, and `EF3` is the case for not trusting a screen alone.

## What each stage establishes

Content review establishes whether items represent the construct as it was
defined: whether the domain is covered, and whether each item belongs where it
was written to belong. It cannot see how an item varies, what it shares with the
others, or how it behaves in different groups.

An empirical screen establishes how the responses behave. It cannot see whether
the domain is covered. A screen would happily keep a scale of ten well-behaved
items that all measured one corner of the construct.

`nomologR` keeps both in the record. A report from a run that started from a
handoff opens with the content review, quoting the panel's decisions in
`contentvalidR`'s words, and each empirical flag carries its own explanation.
Neither is presented as a verdict on the other:

```{r report, eval=FALSE}
nomo_report(run, file = "walkthrough-report.html")
```
