---
title: "From item audit to exploratory structure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From item audit to exploratory structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(nomologR)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Why this workflow is staged

Exploratory factor analysis is not a scale-purification button. Before interpreting
an exploratory factor solution, researchers should know what items are being
analyzed, how they are coded, whether the correlation model matches the intended
measurement level, and how many dimensions are substantively plausible.

`nomologR` therefore separates three questions:

1. `nomo_screen()` — **What does the item/data set look like?**
2. `nomo_factors()` — **How many latent dimensions deserve investigation?**
3. `nomo_efa()` — **What does a requested exploratory structure look like?**

The handoff is deliberate: no stage silently deletes items, changes the requested
factor count, or treats a numerical reference value as proof.

# A reproducible two-factor example

The example below has two correlated latent variables with four indicators each.
The population structure is intentionally clear so that the workflow can be
inspected against a known data-generating process.

```{r simulate}
set.seed(2026)
n <- 400

f1 <- rnorm(n)
f2 <- 0.35 * f1 + sqrt(1 - 0.35^2) * rnorm(n)

dat <- data.frame(
  a1 = .82 * f1 + rnorm(n, sd = .55),
  a2 = .78 * f1 + rnorm(n, sd = .60),
  a3 = .75 * f1 + rnorm(n, sd = .62),
  a4 = .80 * f1 + rnorm(n, sd = .58),
  b1 = .82 * f2 + rnorm(n, sd = .55),
  b2 = .78 * f2 + rnorm(n, sd = .60),
  b3 = .75 * f2 + rnorm(n, sd = .62),
  b4 = .80 * f2 + rnorm(n, sd = .58)
)
```

# Step 1: audit the items

```{r screen}
scr <- nomo_screen(dat)
summary(scr)
```

The screening object is descriptive and diagnostic. A flag is a reason to inspect
an item or coding decision, not an instruction to remove an item.

# Step 2: investigate factor retention

```{r factors}
fac <- nomo_factors(
  dat,
  criterion_set = "core",
  n_iter = 20,
  seed = 2026
)
summary(fac)
```

Parallel analysis is the primary retention suggestion, with MAP and EKC providing
complementary evidence. When methods disagree, the disagreement is part of the
result rather than something `nomologR` resolves with a majority-vote rule.

# Step 3: hand the retention result into EFA

```{r efa}
efa <- nomo_efa(
  dat,
  factors = fac
)
summary(efa)
```

Passing the `nomo_factors` object carries forward the selected item set, modeling
types, correlation model, and missing-data strategy. The primary
parallel-analysis count becomes the **requested** EFA factor count. If M2 identified
neighboring plausible solutions, `nomo_efa()` records that context and recommends
comparing those solutions rather than treating the handoff as proof.

# Read the item table as evidence, not a deletion list

```{r item-table}
efa$item_summary
```

The item table reports:

- the largest pattern loading;
- the second-largest pattern loading;
- the gap between them;
- communality and uniqueness;
- loading complexity;
- `KEEP`, `REVIEW`, or `STRONG REVIEW`;
- a plain-language explanation for every numerical flag.

`KEEP` means that no configured numerical teaching-reference flag fired. It does
**not** mean that theory, wording, content coverage, redundancy, or external
validity evidence has approved the item.

# Pattern versus structure matrices

With oblique rotation, the pattern matrix contains regression-like factor
coefficients and is the primary matrix for interpreting which indicators define
which factors. The structure matrix contains item-factor correlations and also
reflects correlations among the factors.

```{r matrices}
round(efa$pattern_matrix, 2)
round(efa$structure_matrix, 2)
round(efa$factor_correlations, 2)
```

```{r pattern-plot}
plot(efa, type = "pattern")
```

# Residual strain

A plausible loading pattern can still leave localized relationships unexplained.
`nomo_efa()` therefore returns the reproduced correlation matrix, the residual
matrix, the off-diagonal RMSR, and ranked residual pairs.

```{r residuals}
efa$rmsr
head(efa$residual_pairs, 10)
```

```{r residual-plot}
plot(efa, type = "residuals")
```

Residuals are diagnostic evidence. `nomologR` does not search for a better model,
add correlated residuals, change the factor count, or remove indicators behind
the user's back.

# Researcher control remains explicit

The default is common-factor MINRES extraction with oblimin rotation. Researchers
may choose another common-factor extraction or rotation, but those choices remain
visible in the result and decision log. For example:

```{r orthogonal, eval=FALSE}
efa_varimax <- nomo_efa(
  dat,
  factors = 2,
  rotation = "varimax"
)
```

An orthogonal choice is permitted rather than prohibited, but the output asks the
researcher to justify the assumption that the factors should be uncorrelated.

For numeric Likert items, declare the intended modeling level rather than assuming
that integer storage proves continuous or ordinal measurement:

```{r ordinal-example, eval=FALSE}
types <- c(
  q1 = "ordinal",
  q2 = "ordinal",
  q3 = "ordinal",
  q4 = "ordinal"
)

efa_ord <- nomo_efa(
  dat_likert,
  factors = 1,
  types = types
)
```

# What should happen next?

EFA contributes exploratory evidence about latent structure. A defensible
measurement workflow still requires theoretical interpretation and, where
feasible, confirmation on fresh or holdout data. The next `nomologR` milestone is
confirmatory factor analysis.

Methodological treatments that motivate the package's EFA philosophy include
Fabrigar, Wegener, MacCallum, and Strahan (1999), Watkins (2018), and later
reviews emphasizing common-factor models, evidence-based retention, and oblique
rotation when factor correlations are plausible.
