---
title: "From item audit to exploratory structure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From item audit to exploratory structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(nomologR)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Why this workflow is staged

Exploratory factor analysis is not a scale-purification button. Before
interpreting an exploratory factor solution, researchers should know what items
are being analyzed, how they are coded, whether the correlation model matches
the intended measurement level, and how many dimensions are substantively
plausible.

`nomologR` therefore separates three questions:

1. `nomo_screen()` — **What does the item and response data look like?**
2. `nomo_factors()` — **How many latent dimensions deserve investigation?**
3. `nomo_efa()` — **What does a requested exploratory structure look like?**

The handoff is deliberate: no stage silently deletes items, changes the
requested factor count, or treats a numerical reference value as proof.

# A teaching dataset with a known answer

`nomo_demo_continuous` was simulated from a documented population model, so the
package's evidence can be checked against the truth:

- two factors, A and B, correlated .40;
- items `a1`–`a4` and `b1`–`b4` are ordinary indicators (population loadings
  .70 to .80);
- `a5` **cross-loads** on both factors (.45 on A and .35 on B);
- `b5` is a **weak** indicator (.30 on B);
- `a2` and `b3` contain a small amount of missing data.

```{r data}
str(nomo_demo_continuous)
colSums(is.na(nomo_demo_continuous))
```

With real data, of course, nobody hands you the population model. The point of
a known-answer dataset is to learn what the evidence looks like when you *do*
know what is true.

# Step 1: audit the items

```{r screen}
scr <- nomo_screen(nomo_demo_continuous)
summary(scr)
```

The screening object is descriptive and diagnostic. The missing values in `a2`
and `b3` are visible here, before any model decides how to handle them. A flag
is a reason to inspect an item or a coding decision, not an instruction to
remove it.

```{r screen-plot}
plot(scr)
```

## Careless responding

The audit above is about items. A second question is about respondents: did
everyone read the items? The data below is **simulated**: four five-point scales
of six items, two reverse keyed in each, with 15 respondents who gave the same
answer throughout and 15 who answered at random.

```{r effort-data}
set.seed(34)
n <- 400
factors <- replicate(4, rnorm(n))
respond <- function(f) pmin(5, pmax(1, round(3 + f + rnorm(n, sd = .7))))

responses <- list()
reverse <- character()
scales <- list()
for (s in 1:4) {
  items <- paste0("s", s, "_", 1:6)
  for (j in 1:6) {
    r <- respond(factors[, s])
    if (j %in% c(2, 5)) {
      r <- 6 - r
      reverse <- c(reverse, items[j])
    }
    responses[[items[j]]] <- r
  }
  scales[[paste0("S", s)]] <- items
}
survey <- as.data.frame(responses)
survey[1:15, ] <- 3
survey[16:30, ] <- matrix(sample(1:5, 15 * ncol(survey), TRUE), 15, ncol(survey))
```

`effort = TRUE` adds case-level indices from the careless-responding
literature. Reverse keying and the response range are declared rather than
inferred, because a range guessed from the data is wrong whenever a category
went unused:

```{r effort}
careful <- nomo_screen(
  survey,
  effort = TRUE,
  scales = scales,
  reverse = reverse,
  scale_range = c(1, 5)
)
careful
```

The indices do not agree, and they are not supposed to. Compare what two of
them say about the respondents who answered identically throughout:

```{r effort-disagreement}
nomo_table(careful, "effort")[1:3, c(
  "row", "long_string", "inter_item_sd", "flagged_by"
)]
```

Long-string flags every one of them. Inter-item standard deviation gives them
a score of zero, the most consistent possible. That is not a contradiction.
Inter-item standard deviation detects *random* responding (Marjanovic et al.,
2015), and a respondent who never varies is not random. Sorting on it alone
would keep exactly the respondents long-string exists to find, which is why
Curran (2016) recommends using these methods in series.

Two further cautions come from the sources. Huang et al. (2012) found the
indices they recommended identified attentive respondents well and random
responders poorly, so a case with no flag has not been shown to be attentive.
And each antonym, synonym, and even-odd value is a correlation computed across
a handful of pairs or scales; with few of them, attentive respondents cross
zero by chance, and the decision log says so.

Nothing is removed. Deciding what to do with a flagged respondent is a
research decision, made with the design and the data collection in view.

# Step 2: investigate factor retention

```{r factors}
fac <- nomo_factors(
  nomo_demo_continuous,
  criterion_set = "core",
  seed = 2026
)
summary(fac)
```

Parallel analysis is the primary retention evidence, with MAP and the empirical
Kaiser criterion (EKC) providing complementary evidence. Here parallel analysis
and both MAP variants agree on two dimensions, which matches the population
model. When methods disagree, the disagreement is part of the result rather than
something `nomologR` resolves with a majority vote.

Notice that EKC was requested but not run. The default pairwise-complete
correlations use different numbers of cases for different item pairs (because of
the missing values in `a2` and `b3`), and EKC needs one common sample size.
`nomologR` reports that reason instead of silently substituting another
criterion. Choosing `missing = "complete"` would make EKC available, at the cost
of analyzing only the cases with no missing items — a trade-off for the
researcher to make and record.

```{r scree-plot}
plot(fac, type = "scree")
```

## Historical note: eigenvalues greater than one

For decades, many researchers retained every component with an eigenvalue
greater than one (Guttman, 1954; Kaiser, 1960), often as part of the "Little
Jiffy" routine of principal components, eigenvalues greater than one, and
varimax rotation. Simulation research has shown that this rule frequently
retains too many dimensions, and parallel analysis (Horn, 1965) is now preferred.
`nomologR` can display the legacy rule with `criterion_set = "all"`, labeled as
historical context and excluded from the synthesis, so that learners can
recognize it in published work without relying on it.

# Step 3: hand the retention evidence into EFA

```{r efa}
efa <- nomo_efa(
  nomo_demo_continuous,
  factors = fac
)
summary(efa)
```

Passing the `nomo_factors` object carries forward the item set, modeling types,
correlation model, and missing-data strategy. The primary parallel-analysis
count becomes the **requested** EFA factor count. If the retention evidence had
identified neighboring plausible solutions, `nomo_efa()` would record that
context and recommend comparing them rather than treating the handoff as proof.

# Read the item table as evidence, not a deletion list

```{r item-table}
item_view <- efa$item_summary[, c(
  "item", "primary_factor", "primary_loading", "secondary_loading",
  "communality", "attention"
)]
item_view
```

```{r item-values, include=FALSE}
row_of <- function(item) efa$item_summary[efa$item_summary$item == item, ]
a5 <- row_of("a5")
b5 <- row_of("b5")
b4 <- row_of("b4")
```

Compare the table with the known population model:

- `a5` receives **`r a5$attention`**. Its secondary loading of
  `r sprintf("%.2f", a5$secondary_loading)` reaches the cross-loading reference,
  exactly the feature built into the data.
- `b5` receives **`r b5$attention`**. Its primary loading of
  `r sprintf("%.2f", b5$primary_loading)` and communality of
  `r sprintf("%.2f", b5$communality)` reflect the weak population loading.
- `b4` receives **`r b4$attention`** even though its population loading is .70.
  Its sample communality of `r sprintf("%.3f", b4$communality)` sits essentially
  at the .40 reference. This is sampling variability around a teaching
  reference — a reminder that references are prompts, not cliffs.

`KEEP` means that no configured numerical teaching reference fired. It does
**not** mean that theory, wording, content coverage, redundancy, or external
validity evidence has approved the item. Likewise, `REVIEW` does not mean
"delete." Defensible options for `a5` include retaining it because its content
is essential, revising its wording, or evaluating a model without it — and
recording the rationale either way.

# Pattern versus structure matrices

With oblique rotation, the pattern matrix contains regression-like factor
coefficients and is the primary matrix for interpreting which indicators define
which factors. The structure matrix contains item–factor correlations and also
reflects the correlation between factors.

```{r matrices}
round(efa$pattern_matrix, 2)
round(efa$structure_matrix, 2)
round(efa$factor_correlations, 2)
```

```{r pattern-plot}
plot(efa, type = "pattern")
```

# Residual strain

A plausible loading pattern can still leave localized relationships
unexplained. `nomo_efa()` therefore returns the reproduced correlation matrix,
the residual matrix, the off-diagonal RMSR, and ranked residual pairs.

```{r residuals}
efa$rmsr
head(efa$residual_pairs, 10)
```

```{r residual-plot}
plot(efa, type = "residuals")
```

Residuals are diagnostic evidence. `nomologR` does not search for a better
model, add correlated residuals, change the factor count, or remove indicators
behind the user's back.

# Researcher control remains explicit

The default is common-factor MINRES extraction with oblimin rotation.
Researchers may choose another common-factor extraction or rotation, but those
choices remain visible in the result and decision log.

```{r orthogonal}
efa_varimax <- nomo_efa(
  nomo_demo_continuous,
  factors = 2,
  rotation = "varimax"
)
efa_varimax$item_summary[, c("item", "primary_loading", "secondary_loading", "attention")]
```

An orthogonal rotation is permitted rather than prohibited, but it forces the
factors to be uncorrelated even though A and B correlate .40 in the population.
The decision log asks the researcher to justify that assumption.

# Ordered response categories

Many rating scales produce ordered categories rather than continuous scores.
`nomo_demo_ordinal` stores the same latent responses as five-category ordered
factors, so `nomologR` uses polychoric correlations automatically:

```{r ordinal-factors}
fac_ord <- nomo_factors(nomo_demo_ordinal, seed = 2026)
fac_ord$correlation

efa_ord <- nomo_efa(nomo_demo_ordinal, factors = fac_ord)
efa_ord$item_summary[, c("item", "primary_loading", "secondary_loading", "attention")]
```

When Likert-type responses are stored as integers instead of ordered factors,
integer storage alone does not tell `nomologR` whether the researcher intends a
continuous or ordinal model. Declare the modeling level explicitly. This chunk
only illustrates the syntax, because the teaching data already store ordered
factors:

```{r ordinal-types, eval=FALSE}
efa_declared <- nomo_efa(
  my_likert_items,
  factors = 2,
  types = c(q1 = "ordinal", q2 = "ordinal", q3 = "ordinal", q4 = "ordinal")
)
```

# What should happen next?

EFA contributes exploratory evidence about latent structure. A defensible
measurement workflow still requires theoretical interpretation and, where
feasible, confirmation on fresh or holdout data. Continue with
[From CFA to a defensible measurement model](measurement-model-evidence.html).

# Research basis

The package's EFA philosophy follows reviews that documented problematic
historical defaults — principal components presented as factor analysis,
eigenvalues greater than one, orthogonal rotation, and mechanical deletion — and
recommended common-factor models, evidence-based retention, and oblique rotation
when factors may correlate (Fabrigar et al., 1999; Conway & Huffcutt, 2003;
Browne, 2001; Watkins, 2018). Parallel analysis originates with Horn (1965); the
eigenvalue-greater-than-one rule with Guttman (1954) and Kaiser (1960). See the
[research basis](research-basis.html) article for full references and the
broader historical-to-contemporary map.
