---
title: "Guided workflow with nomo_run()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guided workflow with nomo_run()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(nomologR)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

`nomo_run()` is an orchestrator, not an automatic scale-decider.

Its job is to make the staged workflow easier to reproduce while preserving the
package's central rule:

> Evidence may flow automatically. Consequential decisions do not.

This walkthrough uses `nomo_demo_network`, a simulated validation study with
three constructs (Agency, Persistence, and Social desirability), an observed
Performance outcome, and two administration groups. See `?nomo_demo_network`.

## A fresh run

```{r fresh}
scales <- list(
  Agency = c("ag1", "ag2", "ag3", "ag4"),
  Persistence = c("pe1", "pe2", "pe3", "pe4"),
  SocialDesirability = c("sd1", "sd2", "sd3")
)

run <- nomo_run(
  data = nomo_demo_network,
  scales = scales,
  settings = list(factors = list(seed = 2026))
)

run
```

The initial run audits each scale's items and computes factor-retention
evidence. It then pauses before EFA. The factor count is not silently copied
from parallel analysis.

```{r requests-1}
nomo_table(run, "requests")
```

Each decision request separates:

- **Observation** — what the evidence shows;
- **Reason** — why a researcher decision is required;
- **Options** — what can defensibly happen next;
- **Consequence** — what the choice changes.

## Resume with an explicit factor count

```{r factor-count}
run <- nomo_run(
  resume = run,
  decisions = list(
    factor_count = list(
      value = c(Agency = 1, Persistence = 1, SocialDesirability = 1),
      rationale = paste(
        "Each scale was written to measure one construct, and the",
        "retention evidence was reviewed before choosing one factor."
      )
    )
  )
)

run
```

The completed screening and factor-retention objects are reused. EFA is fitted
using the explicit factor-count decision, with the retention stage's correlation
and modeling context retained. The workflow then pauses before CFA model
specification.

## Confirmatory model handoff

```{r cfa-model}
run <- nomo_run(
  resume = run,
  decisions = list(
    cfa_model = list(
      value = nomo_model(scales),
      rationale = "Prespecified three-construct measurement model."
    )
  )
)

run
```

`nomo_run()` does not translate the EFA into CFA syntax. The researcher owns the
confirmatory model; `nomo_model()` simply writes the simple-structure syntax for
the scales that were already declared.

Once supplied, the pipeline computes:

```text
nomo_cfa()
  |
  +-- nomo_reliability()
  |
  +-- nomo_validity()
```

and pauses again.

## Why pause after measurement evidence?

A model can converge and still contain evidence that deserves review.
Reliability or AVE can look strong while construct separation or local CFA
evidence remains strained. Therefore the pipeline does not automatically carry
the model into invariance or structural theory tests.

```{r requests-2}
nomo_table(run, "requests")
```

The researcher chooses either `measurement_model = "proceed"` or
`measurement_model = "revise"`. Choosing `revise` does not trigger an automated
modification-index search, parameter freeing, or item deletion; the current
workflow remains an auditable record. To carry a revision forward with its
reasoning attached, use `nomo_revise()`, shown in
[Revising with a recorded lineage](#revising-with-a-recorded-lineage) below.

## Optional invariance and network branches

Future-stage settings can be added at a pause without recomputing completed
stages.

```{r branches}
h <- nomo_hypotheses(
  "Agency -> Persistence" = positive(min = .20),
  "Agency <-> SocialDesirability" = negligible(within = c(-.15, .15)),
  "Agency -> Performance" = positive()
)

run <- nomo_run(
  resume = run,
  settings = list(
    invariance = list(
      group = "group",
      levels = c("configural", "metric", "scalar"),
      localize = TRUE
    ),
    network = list(
      hypotheses = h
    )
  )
)
```

Then continue explicitly:

```{r proceed}
run <- nomo_run(
  resume = run,
  decisions = list(
    measurement_model = list(
      value = "proceed",
      rationale = "Measurement evidence reviewed before planned downstream tests."
    )
  )
)

run
nomo_table(run, "stages")
```

The completed component objects remain available for direct inspection:

```{r components}
nomo_table(run$results$invariance, "fit")[, c("level", "cfi", "rmsea", "delta_cfi")]

nomo_table(run$results$network, "hypotheses")[, c(
  "relation", "estimate", "ci_lower", "ci_upper", "concordance"
)]
```

The invariance branch retains its diagnostics without automatically releasing
constraints (see [Measurement invariance](measurement-invariance.html) for the
`ag3` intercept). The network branch uses explicit hypotheses and does not
create a single nomological-validity score.

## Prespecified one-call workflow

Researchers who have already made the consequential decisions — for example in
a preregistration — can supply them together. `mode = "research"` gives a more
compact presentation of the same statistical behavior:

```{r one-call}
run_prespecified <- nomo_run(
  data = nomo_demo_network,
  scales = scales,
  mode = "research",
  decisions = list(
    factor_count = c(Agency = 1, Persistence = 1, SocialDesirability = 1),
    cfa_model = nomo_model(scales),
    measurement_model = "proceed"
  ),
  settings = list(
    factors = list(seed = 2026),
    network = list(hypotheses = h)
  )
)

run_prespecified
```

This is still not hidden automation: every consequential decision was explicit
before the run began.

## Revising with a recorded lineage

Suppose the measurement evidence prompts a change. `nomo_revise()` creates a
child workflow that keeps the parent as its documented ancestor, instead of
starting an unrelated run:

```{r revise}
revised <- nomo_revise(
  run,
  cfa_model = paste(
    "Agency =~ ag1 + ag2 + ag3 + ag4",
    "Persistence =~ pe1 + pe2 + pe3 + pe4",
    "SocialDesirability =~ sd1 + sd2 + sd3",
    "ag1 ~~ ag2",
    sep = "\n"
  ),
  rationale = paste(
    "Items ag1 and ag2 use nearly identical wording, so their residual",
    "association plausibly exceeds what the common factor explains."
  ),
  origin = "post_hoc"
)

revised
```

The child reruns the staged evidence with the revised model and pauses at the
measurement review, so the revised evidence is inspected before any downstream
branch runs. The parent object is unchanged and remains a complete record.

```{r revise-lineage}
nomo_table(revised, "lineage")[, c(
  "revision", "change_type", "origin", "rationale", "comparison"
)]
```

Because a revision is a claim that one model is preferable to another,
`nomo_revise()` compares them with `nomo_compare()`:

```{r revise-comparison}
nomo_table(revised$revision_comparison, "comparisons")[, c(
  "model", "relation", "chisq_diff", "df_diff", "p_value", "delta_cfi"
)]
```

```{r revise-values, include=FALSE}
rev_cmp <- revised$revision_comparison$comparisons
rev_log <- revised$decision_log[revised$decision_log$id == "revision", ]
fmt_rev_p <- if (rev_cmp$p_value < .001) {
  "< .001"
} else {
  paste("=", formatC(rev_cmp$p_value, format = "f", digits = 3))
}
```

Here the revision changes almost nothing (chi-square difference =
`r sprintf("%.2f", rev_cmp$chisq_diff)`, df = `r rev_cmp$df_diff`,
p `r fmt_rev_p`; CFI changes by `r sprintf("%+.4f", rev_cmp$delta_cfi)`). Once
the common factor is accounted for, the residual association these two items
were expected to share is close to zero, so the data give no reason to prefer
the revised model over its parent. A revision is not automatically an
improvement, and `nomo_revise()` does not choose: it records the attempt, its
rationale, and the comparison, so keeping the parent model is as visible as
adopting the revision.

The origin still matters. The change was prompted by inspecting these data, so
the decision log records it as post hoc and says what follows from that:

```{r revise-log}
cat(rev_log$consequence)
```

Revisions chain. A revision of `revised` would carry two lineage rows, so the
full path from the original model to the reported one stays visible in
`nomo_table(x, "lineage")` and in the report's revision-lineage section.

## Starting from content review

Scale development begins before response data exist. The companion package
[`contentvalidR`](https://github.com/JUhalt/contentvalidR) handles content
review: expert relevance panels, item sorts, and Delphi rounds. It ends in a
*handoff*, which records the items carried forward, each held-back item with its
reasons, and the rule that made each decision. `nomologR` takes the handoff
itself, not a copied list of item names, so the reasons travel with the items.

In `contentvalidR`, a handoff is made from a fitted review. The chunk below is
shown, not run:

```r
fit <- contentvalidR::sort_validity(sorts)
handoff <- contentvalidR::content_handoff(
  fit,
  reverse_keyed = c("EF2", "TF2"),
  response_scale = c(1, 5)
)
```

`nomologR` does not depend on `contentvalidR`. This package ships the handoff
that `contentvalidR` 0.7.0 produces for its walkthrough item sort, so the rest of
this section runs without it:

```{r handoff}
handoff <- readRDS(
  system.file("extdata", "content-handoff-walkthrough.rds", package = "nomologR")
)
handoff$items
```

Responses to those items, simulated here, with EF2 and TF2 worded in reverse:

```{r handoff-data}
set.seed(46)
n <- 400
ef <- rnorm(n)
tf <- 0.4 * ef + sqrt(1 - 0.4^2) * rnorm(n)
likert <- function(f) pmin(5, pmax(1, round(3 + 0.8 * f + rnorm(n, sd = 0.8))))
responses <- data.frame(
  sapply(paste0("EF", 1:6), function(i) likert(ef)),
  sapply(paste0("TF", 1:6), function(i) likert(tf))
)
responses$EF2 <- 6 - responses$EF2
responses$TF2 <- 6 - responses$TF2
```

Passing the handoff as `items` screens only the carried items:

```{r handoff-screen}
screened <- nomo_screen(responses, items = handoff, effort = TRUE)
screened$items

review_log <- nomo_table(screened, "decision_log")
cat(review_log$observation[review_log$object %in% c("content_review", "EF5", "TF5")],
    sep = "\n")
```

EF5 and TF5 were held back by the item sort. They are not analyzed, and nothing
here reinstates them. The log quotes their status and recommendation in
`contentvalidR`'s own words. The keying declared at content review reached the
careless-responding indices without being retyped. An item is never treated as
forward keyed because keying was undeclared, and a response scale is never
inferred from the data.

Passed as `scales`, the handoff defines a guided run. Its design log records
that the item membership came from content review, not from these data:

```{r handoff-run}
run_reviewed <- nomo_run(
  responses,
  scales = handoff,
  settings = list(factors = list(seed = 46))
)
run_reviewed$scales

design <- run_reviewed$decision_log
design[design$source == "content_review", c("id", "scope")]
```

A report from this run opens with a content-review section, so the archive
starts where the validity argument starts.

Some reviews have no construct mapping, such as an expert relevance panel
rating a single item set. A handoff from one of those can be screened, but a
guided run refuses it. The run needs scales, and `nomologR` does not invent
construct membership. A handoff from a later `contentvalidR` release, with a
schema version this release does not read, is refused with both package
versions named.

## Reproducibility and provenance

```{r provenance}
nomo_table(run, "decisions")
nomo_table(run, "settings")
nomo_table(run, "recipe")
head(nomo_table(run, "component_log"), 10)
```

The guided object retains:

- source sample roles;
- supplied scale membership;
- component settings;
- explicit decisions and rationales;
- complete component result objects;
- outstanding decision requests;
- component decision and evidence logs;
- the sequence of resume calls.

This structure feeds the reproducible report described in
[Archiving a nomologR workflow](reproducible-report.html).

## Research basis

Staged scale development with explicit decisions follows guidance such as Clark
and Watson (1995, 2019), Hinkin (1998), and Boateng et al. (2018). Recording
decisions and rationales responds to evidence that undisclosed analytic
flexibility undermines measurement claims (Simmons, Nelson, & Simonsohn, 2011;
Flake, Pek, & Hehman, 2017; Flake & Fried, 2020). Full references are in
`?nomo_run` and the [research basis](research-basis.html) article.
