---
title: "Guided workflow with nomo_run()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guided workflow with nomo_run()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(nomologR)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

`nomo_run()` is an orchestrator, not an automatic scale-decider.

Its job is to make the staged workflow easier to reproduce while preserving the
package's central rule:

> Evidence may flow automatically. Consequential decisions do not.

This walkthrough uses `nomo_demo_network`, a simulated validation study with
three constructs (Agency, Persistence, and Social desirability), an observed
Performance outcome, and two administration groups. See `?nomo_demo_network`.

## A fresh run

```{r fresh}
scales <- list(
  Agency = c("ag1", "ag2", "ag3", "ag4"),
  Persistence = c("pe1", "pe2", "pe3", "pe4"),
  SocialDesirability = c("sd1", "sd2", "sd3")
)

run <- nomo_run(
  data = nomo_demo_network,
  scales = scales,
  settings = list(factors = list(seed = 2026))
)

run
```

The initial run audits each scale's items and computes factor-retention
evidence. It then pauses before EFA. The factor count is not silently copied
from parallel analysis.

```{r requests-1}
nomo_table(run, "requests")
```

Each decision request separates:

- **Observation** — what the evidence shows;
- **Reason** — why a researcher decision is required;
- **Options** — what can defensibly happen next;
- **Consequence** — what the choice changes.

## Resume with an explicit factor count

```{r factor-count}
run <- nomo_run(
  resume = run,
  decisions = list(
    factor_count = list(
      value = c(Agency = 1, Persistence = 1, SocialDesirability = 1),
      rationale = paste(
        "Each scale was written to measure one construct, and the",
        "retention evidence was reviewed before choosing one factor."
      )
    )
  )
)

run
```

The completed screening and factor-retention objects are reused. EFA is fitted
using the explicit factor-count decision, with the retention stage's correlation
and modeling context retained. The workflow then pauses before CFA model
specification.

## Confirmatory model handoff

```{r cfa-model}
run <- nomo_run(
  resume = run,
  decisions = list(
    cfa_model = list(
      value = nomo_model(scales),
      rationale = "Prespecified three-construct measurement model."
    )
  )
)

run
```

`nomo_run()` does not translate the EFA into CFA syntax. The researcher owns the
confirmatory model; `nomo_model()` simply writes the simple-structure syntax for
the scales that were already declared.

Once supplied, the pipeline computes:

```text
nomo_cfa()
  |
  +-- nomo_reliability()
  |
  +-- nomo_validity()
```

and pauses again.

## Why pause after measurement evidence?

A model can converge and still contain evidence that deserves review.
Reliability or AVE can look strong while construct separation or local CFA
evidence remains strained. Therefore the pipeline does not automatically carry
the model into invariance or structural theory tests.

```{r requests-2}
nomo_table(run, "requests")
```

The researcher chooses either `measurement_model = "proceed"` or
`measurement_model = "revise"`. Choosing `revise` does not trigger an automated
modification-index search, parameter freeing, or item deletion; the current
workflow remains an auditable record and the revised analysis starts as a new
run. Linking a revised run to its parent with a formal model comparison is
planned ([#27](https://github.com/JUhalt/nomologR/issues/27),
[#28](https://github.com/JUhalt/nomologR/issues/28)).

## Optional invariance and network branches

Future-stage settings can be added at a pause without recomputing completed
stages.

```{r branches}
h <- nomo_hypotheses(
  "Agency -> Persistence" = positive(min = .20),
  "Agency <-> SocialDesirability" = negligible(within = c(-.15, .15)),
  "Agency -> Performance" = positive()
)

run <- nomo_run(
  resume = run,
  settings = list(
    invariance = list(
      group = "group",
      levels = c("configural", "metric", "scalar"),
      localize = TRUE
    ),
    network = list(
      hypotheses = h
    )
  )
)
```

Then continue explicitly:

```{r proceed}
run <- nomo_run(
  resume = run,
  decisions = list(
    measurement_model = list(
      value = "proceed",
      rationale = "Measurement evidence reviewed before planned downstream tests."
    )
  )
)

run
nomo_table(run, "stages")
```

The completed component objects remain available for direct inspection:

```{r components}
nomo_table(run$results$invariance, "fit")[, c("level", "cfi", "rmsea", "delta_cfi")]

nomo_table(run$results$network, "hypotheses")[, c(
  "relation", "estimate", "ci_lower", "ci_upper", "concordance"
)]
```

The invariance branch retains its diagnostics without automatically releasing
constraints (see [Measurement invariance](measurement-invariance.html) for the
`ag3` intercept). The network branch uses explicit hypotheses and does not
create a single nomological-validity score.

## Prespecified one-call workflow

Researchers who have already made the consequential decisions — for example in
a preregistration — can supply them together. `mode = "research"` gives a more
compact presentation of the same statistical behavior:

```{r one-call}
run_prespecified <- nomo_run(
  data = nomo_demo_network,
  scales = scales,
  mode = "research",
  decisions = list(
    factor_count = c(Agency = 1, Persistence = 1, SocialDesirability = 1),
    cfa_model = nomo_model(scales),
    measurement_model = "proceed"
  ),
  settings = list(
    factors = list(seed = 2026),
    network = list(hypotheses = h)
  )
)

run_prespecified
```

This is still not hidden automation: every consequential decision was explicit
before the run began.

## Reproducibility and provenance

```{r provenance}
nomo_table(run, "decisions")
nomo_table(run, "settings")
nomo_table(run, "recipe")
head(nomo_table(run, "component_log"), 10)
```

The guided object retains:

- source sample roles;
- supplied scale membership;
- component settings;
- explicit decisions and rationales;
- complete component result objects;
- outstanding decision requests;
- component decision and evidence logs;
- the sequence of resume calls.

This structure feeds the reproducible report described in
[Archiving a nomologR workflow](reproducible-report.html).

## Research basis

Staged scale development with explicit decisions follows guidance such as Clark
and Watson (1995, 2019), Hinkin (1998), and Boateng et al. (2018). Recording
decisions and rationales responds to evidence that undisclosed analytic
flexibility undermines measurement claims (Simmons, Nelson, & Simonsohn, 2011;
Flake, Pek, & Hehman, 2017; Flake & Fried, 2020). Full references are in
`?nomo_run` and the [research basis](research-basis.html) article.
