---
title: "Total and subscale scores: bifactor and higher-order models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Total and subscale scores: bifactor and higher-order models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(nomologR)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4
)
```

Many published scales report a **total score** and several **subscale
scores**. A correlated-factors CFA can show that the subscales are
distinguishable, but it cannot say whether a total score is defensible, or how
much a subscale score adds once the total is known. Those questions need a
model with a general factor and indices that divide each score's variance
between the general factor and the subscales.

This article shows how `nomologR` fits and evaluates the two common models for
that purpose, and why model fit alone cannot choose between them.

# Two ways to model a general factor

- A **higher-order** model has first-order factors, one per subscale, whose
  correlations are explained by a second-order factor. The general factor
  reaches the items only through the first-order factors.
- A **bifactor** model has a general factor measured directly by every item,
  plus a group factor for each subscale. All factors are orthogonal, so each
  item's variance divides cleanly between the general factor, its group
  factor, and error.

The bifactor model dates to Holzinger and Swineford (1937), and Schmid and
Leiman (1957) showed how to re-express a higher-order solution in bifactor
form. Bifactor models were later rediscovered as a way to evaluate
multidimensional measures (Reise, 2012), with indices for judging total and
subscale scores (Reise, Bonifay, & Haviland, 2013; Rodriguez, Reise, &
Haviland, 2016). The [research basis](research-basis.html) article places these
methods in context.

# An example with a general factor

The data below are simulated from a bifactor population: twelve items in three
four-item subscales, with a general factor that every item reflects and a
weaker group factor for each subscale.

```{r simulate}
simulate_from <- function(sigma, n, seed) {
  set.seed(seed)
  z <- matrix(rnorm(n * ncol(sigma)), n, ncol(sigma))
  dat <- as.data.frame(z %*% chol(sigma))
  names(dat) <- colnames(sigma)
  dat
}

items <- paste0("x", 1:12)
subscale <- rep(1:3, each = 4)
general_loading <- c(.70, .60, .65, .55, .60, .50, .70, .60, .50, .60, .55, .65)
group_loading <- c(.40, .50, .30, .45, .50, .35, .40, .30, .45, .40, .50, .35)

L <- cbind(
  general_loading,
  sapply(1:3, function(k) ifelse(subscale == k, group_loading, 0))
)
sigma_bifactor <- L %*% t(L)
diag(sigma_bifactor) <- 1
dimnames(sigma_bifactor) <- list(items, items)

dat <- simulate_from(sigma_bifactor, n = 600, seed = 2026)
```

`nomo_model()` writes the syntax for all three structures from the same list of
subscales. The researcher chooses the structure; `nomologR` never derives one
from exploratory results.

```{r models}
subscales <- list(
  Focus = paste0("x", 1:4),
  Drive = paste0("x", 5:8),
  Poise = paste0("x", 9:12)
)

nomo_model(subscales, structure = "bifactor", general = "G")
```

The bifactor syntax states its identification explicitly: each factor's first
loading is freed, each factor's variance is fixed to 1, and every covariance
among factors is fixed to 0. A higher-order model with three first-order factors
carries a note that matters later:

```{r higher-order-syntax}
nomo_model(subscales, structure = "higher_order", general = "G")
```

```{r fit}
correlated <- nomo_cfa(nomo_model(subscales), data = dat)
higher <- nomo_cfa(nomo_model(subscales, "higher_order"), data = dat)
bifactor <- nomo_cfa(nomo_model(subscales, "bifactor"), data = dat)
```

# How much of each score is general?

`nomo_hierarchical()` reads the structure from the fitted model and reports the
indices:

```{r hierarchical}
h <- nomo_hierarchical(bifactor)
h
```

```{r values, include=FALSE}
idx <- function(x, name) x$indices$estimate[x$indices$index == name]
f2 <- function(x) sprintf("%.2f", x)
sub <- h$subscales
hs <- function(name) f2(sub$omega_hierarchical_subscale[sub$subscale == name])
os <- function(name) f2(sub$omega_subscale[sub$subscale == name])
rel <- function(name) f2(sub$omega_hs_relative[sub$subscale == name])
```

Reading the total score:

- **Omega total** (`r f2(idx(h, "omega_total"))`) is the proportion of
  total-score variance explained by all common factors.
- **Omega hierarchical** (`r f2(idx(h, "omega_hierarchical"))`) is the part
  explained by the general factor alone. It is the index that speaks to
  interpreting the total score as a measure of one construct.
- Their ratio (`r f2(idx(h, "omega_hierarchical_relative"))`) says how much of
  the total score's reliable variance is general.
- **ECV** (`r f2(idx(h, "ecv"))`) is the share of common item variance the
  general factor explains, and **PUC** (`r f2(idx(h, "puc"))`) is the share of
  item correlations influenced only by the general factor.

The subscales tell a different story:

```{r subscales}
nomo_table(h, "subscales")[, c(
  "subscale", "omega_subscale", "omega_hierarchical_subscale"
)]
```

Focus has an omega of `r os("Focus")`, which looks like a reliable subscale.
But its **omega hierarchical subscale** is `r hs("Focus")`: once the general
factor is removed, that is the reliable variance specific to Focus. Only
`r rel("Focus")` of a Focus score's reliable variance is specific to Focus; the
rest is the general factor, which the total score already measures. Reise,
Bonifay, and Haviland (2013) frame exactly this question: whether subscale
scores are justified, and how much reliable variance they add after
controlling for the general factor.

The same division is shown for every composite:

```{r variance-plot}
plot(h)
```

`nomologR` does not turn these indices into verdicts. Published heuristic
values exist, but Reise (2012) notes that no benchmark value of ECV establishes
when a general factor is strong enough. Whether a subscale score is worth
reporting depends on its intended use, and the indices are the evidence for
that decision rather than the decision itself.

# Can you score the factors?

The omega indices describe unit-weighted composites: the sum score a researcher
would actually compute. A different question is how well each *factor* can be
recovered at all. The `factors` table answers it two ways:

```{r factor-scores}
nomo_table(h, "factors")
```

`factor_determinacy` is the correlation between a factor and its estimated
factor score (Beauducel, 2011; Rodriguez, Reise, & Haviland, 2016). Gorsuch
(1983) recommended using factor score estimates only above .90. `min_competing_r`
is the lowest correlation two equally valid sets of scores could have. When it
is negative, two researchers scoring the same data with equally defensible
methods could rank people in opposite orders, and both would be consistent with
the model.

`construct_replicability`, Hancock and Mueller's (2001) H, is the proportion of
variance in a factor its own indicators could explain if optimally weighted.
They proposed .70 as a standard.

The two indices answer different questions, and the difference is easy to miss.
They are the same quantity when a construct is unidimensional. Under a bifactor
model they are not: determinacy is computed from the whole reproduced
correlation matrix, so a group factor's score can borrow the other items to
partial out the general factor, while H sees only that factor's own loadings
and treats the rest of each item as uncorrelated residual. Rodriguez et al.
(2016) state that the two can differ here and decline to prefer either, so
`nomologR` reports both and names what each one measures.

Neither threshold is applied as a rule. They are reported as their authors'
recommendations where a value falls below them, in the same way fixed fit-index
cutoffs are treated elsewhere in the package.

# Comparing structures

The three models can be compared with `nomo_compare()`, which requires a
rationale and never selects a model:

```{r compare}
cmp <- nomo_compare(
  correlated, higher, bifactor,
  rationale = "Competing structures for the same twelve items.",
  origin = "a_priori"
)

comparisons <- nomo_table(cmp, "comparisons")[, c(
  "model", "relation", "chisq_diff", "df_diff", "delta_cfi"
)]
comparisons$delta_cfi <- round(comparisons$delta_cfi, 4)
comparisons
```

```{r compare-values, include=FALSE}
chisq <- function(x) unname(lavaan::fitMeasures(x$fit, "chisq"))
dfree <- function(x) unname(lavaan::fitMeasures(x$fit, "df"))
```

Two features of this table are structural rather than empirical:

- The higher-order model is **equivalent** to the correlated-factors model:
  both have chi-square = `r sprintf("%.2f", chisq(higher))` on
  `r dfree(higher)` degrees of freedom. With three first-order factors, the
  second-order part is just identified, so it imposes no testable constraint.
  Fit cannot distinguish the two, which is what the note on the syntax warned.
- The bifactor model is **less constrained**. A higher-order model is a
  constrained version of the bifactor model (Yung, Thissen, & McLeod, 1999), so
  the bifactor model can only fit as well or better.

# Why fit cannot choose the structure

That last point has a consequence. The next data set is simulated from a
**higher-order** population, so the higher-order model is the true one:

```{r higher-order-population}
first_order <- c(.70, .65, .60, .75, .70, .60, .65, .70, .55, .70, .65, .60)
second_order <- c(.80, .70, .60)

F <- sapply(1:3, function(k) ifelse(subscale == k, first_order, 0))
phi <- second_order %*% t(second_order)
diag(phi) <- 1
sigma_higher <- F %*% phi %*% t(F)
diag(sigma_higher) <- 1
dimnames(sigma_higher) <- list(items, items)

dat_higher <- simulate_from(sigma_higher, n = 600, seed = 2027)

true_model <- nomo_cfa(nomo_model(subscales, "higher_order"), data = dat_higher)
bifactor_model <- nomo_cfa(nomo_model(subscales, "bifactor"), data = dat_higher)
```

The higher-order model generated these data, yet the bifactor model fits better:
chi-square = `r sprintf("%.2f", chisq(bifactor_model))` on
`r dfree(bifactor_model)` degrees of freedom, against
`r sprintf("%.2f", chisq(true_model))` on `r dfree(true_model)` for the true
model. That is not a sign that the bifactor model is correct. Its extra
parameters absorb sampling noise, and it will usually fit at least as well as
correlated-factors or higher-order models of the same items (Reise, 2012).

So the choice between structures has to rest on theory: is there a construct
that every item measures directly, with subscale content as a secondary
source? Or is the general construct what the subscales have in common? Fit
comparison can reveal a structure that clearly does not work, but it cannot
make this choice.

# Higher-order models and Schmid-Leiman

`nomo_hierarchical()` also evaluates higher-order models. There the group
sources are the first-order factors' disturbances, and the item loadings are
the Schmid-Leiman decomposition of the higher-order solution, which is exact
for a confirmatory model:

```{r higher-order-indices}
h_higher <- nomo_hierarchical(true_model)
nomo_table(h_higher, "indices")[, c("index", "estimate")]
head(nomo_table(h_higher, "loadings"), 4)
```

Each item's general loading is its first-order loading multiplied by its
factor's second-order loading. Every item in a subscale therefore has the same
ratio of general to group loading, which is why `item_ecv` is identical for all
four Focus items. That proportionality is the constraint that distinguishes a
higher-order model from a bifactor model, where each item's split between
general and group variance is free.

# Ordered indicators

When indicators are declared ordered, `nomo_cfa()` fits the model with WLSMV and
`nomo_hierarchical()` computes the indices on the latent-response scale. Those
omegas are upper bounds on the reliability of the observed ordinal sum score,
and the output says so. The observed ordinal-scale versions are not provided.

# What nomologR does not do

- It does not generate a bifactor or higher-order structure from exploratory
  results. The researcher specifies the structure.
- It does not apply pass/fail thresholds to omega hierarchical, ECV, or PUC.
- It does not choose between correlated-factors, higher-order, and bifactor
  models, and it discloses that fit comparisons favor the bifactor model.
- It evaluates single-group, single-level models, and refuses models whose
  general and group factors are allowed to correlate, because variance cannot
  then be attributed to a source.

Every method used here is listed, with references, by `nomo_methods()`:

```{r methods}
nomo_methods(h)[, c("method", "lineage", "role")]
```

# References

Holzinger, K. J., & Swineford, F. (1937). The bi-factor method.
*Psychometrika, 2*(1), 41–54. https://doi.org/10.1007/BF02287965

Reise, S. P. (2012). The rediscovery of bifactor measurement models.
*Multivariate Behavioral Research, 47*(5), 667–696.
https://doi.org/10.1080/00273171.2012.715555

Reise, S. P., Bonifay, W. E., & Haviland, M. G. (2013). Scoring and modeling
psychological measures in the presence of multidimensionality. *Journal of
Personality Assessment, 95*(2), 129–140.
https://doi.org/10.1080/00223891.2012.725437

Rodriguez, A., Reise, S. P., & Haviland, M. G. (2016). Evaluating bifactor
models: Calculating and interpreting statistical indices. *Psychological
Methods, 21*(2), 137–150. https://doi.org/10.1037/met0000045

Schmid, J., & Leiman, J. M. (1957). The development of hierarchical factor
solutions. *Psychometrika, 22*(1), 53–61. https://doi.org/10.1007/BF02289209

Yung, Y.-F., Thissen, D., & McLeod, L. D. (1999). On the relationship between
the higher-order factor model and the hierarchical factor model.
*Psychometrika, 64*(2), 113–128. https://doi.org/10.1007/BF02294531
