---
title: "From CFA to a defensible measurement model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From CFA to a defensible measurement model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(nomologR)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.8
)
```

# Why measurement evidence is staged

A confirmatory factor model, a reliability coefficient, and a discriminant-validity
statistic answer different questions. Treating them as interchangeable checkboxes
can produce misleading conclusions. `nomologR` therefore keeps the workflow staged:

1. `nomo_cfa()` — **Does the proposed measurement structure reproduce the data reasonably, and where is there model strain?**
2. `nomo_reliability()` — **How consistently does the intended score represent its target under that measurement model?**
3. `nomo_validity()` — **Do indicators converge on their intended constructs, and are theoretically distinct constructs empirically distinguishable?**

The sequence is evidence-guided rather than pass/fail. A reference value can trigger
inspection, but it never automatically deletes an indicator, merges constructs, or
establishes construct validity.

# A healthy two-factor model

We begin with a known two-factor population in which both factors are well measured
and moderately correlated.

```{r healthy-data}
set.seed(2026)
n <- 500

f1 <- rnorm(n)
f2 <- .30 * f1 + sqrt(1 - .30^2) * rnorm(n)

healthy <- data.frame(
  A1 = .82 * f1 + rnorm(n, sd = .55),
  A2 = .79 * f1 + rnorm(n, sd = .58),
  A3 = .76 * f1 + rnorm(n, sd = .62),
  A4 = .80 * f1 + rnorm(n, sd = .57),
  B1 = .83 * f2 + rnorm(n, sd = .54),
  B2 = .78 * f2 + rnorm(n, sd = .60),
  B3 = .75 * f2 + rnorm(n, sd = .63),
  B4 = .81 * f2 + rnorm(n, sd = .56)
)

model <- '
  F1 =~ A1 + A2 + A3 + A4
  F2 =~ B1 + B2 + B3 + B4
'
```

```{r healthy-fit}
cfa <- nomo_cfa(model, data = healthy)

# Modest resampling here keeps vignette build time reasonable.
# Use more resamples for final reporting.
rel <- nomo_reliability(
  cfa,
  ci = "bootstrap",
  ci_boot = 100,
  ci_seed = 2026
)
val <- nomo_validity(cfa, htmt = "both")
```

The compact summaries are intentionally different from the raw engine output. They
surface the evidence most relevant to the user's next decision without repeating the
same full loading table in several places.

```{r healthy-tables}
summary(rel)
summary(val)
```

The figures also avoid duplication. Item-level loading evidence belongs to the CFA;
reliability gets a coefficient plot; convergent validity gets an AVE plot; and
construct separation gets an HTMT-family plot.

```{r healthy-loading-plot}
plot(cfa, type = "loadings")
```

```{r healthy-reliability-plot}
plot(rel)
```

```{r healthy-ave-plot}
plot(val, type = "ave")
```

```{r healthy-discriminant-plot}
plot(val, type = "discriminant")
```

A healthy result contributes several complementary pieces of evidence. It does not
turn the measurement model into a permanently "validated" object. Validity remains
an accumulating argument tied to construct interpretation, population, setting, and
intended use.

# Good global fit does not guarantee good measurement

This is a critical teaching case. The following one-factor model is correctly
specified, but its indicators are weak measures of the latent variable.

```{r weak-data}
set.seed(2027)
n <- 650
f <- rnorm(n)

weak <- data.frame(
  W1 = .30 * f + rnorm(n, sd = .95),
  W2 = .34 * f + rnorm(n, sd = .94),
  W3 = .28 * f + rnorm(n, sd = .96),
  W4 = .32 * f + rnorm(n, sd = .95),
  W5 = .31 * f + rnorm(n, sd = .95)
)

weak_cfa <- nomo_cfa(
  'Weak =~ W1 + W2 + W3 + W4 + W5',
  data = weak
)
weak_rel <- nomo_reliability(weak_cfa)
weak_val <- nomo_validity(weak_cfa, htmt = "none")
```

```{r weak-fit}
weak_cfa$fit_evidence
```

Because the one-factor structure is the correct covariance model, global fit can look
quite good. That does **not** imply that the items strongly represent the construct or
that the resulting score is precise.

```{r weak-measurement}
summary(weak_rel)
summary(weak_val)
```

This is why `nomologR` does not collapse measurement quality into a single global-fit
verdict. Weak standardized loadings, low AVE, and weak model-based reliability remain
important even when CFI, RMSEA, or SRMR look favorable.

# Strong reliability does not guarantee construct separation

The reverse problem is also possible. Two constructs can each be measured very
consistently while remaining difficult to distinguish empirically.

```{r overlap-data}
set.seed(2028)
n <- 700
f1 <- rnorm(n)
f2 <- .94 * f1 + sqrt(1 - .94^2) * rnorm(n)

overlap <- data.frame(
  A1 = .86 * f1 + rnorm(n, sd = .48),
  A2 = .83 * f1 + rnorm(n, sd = .51),
  A3 = .84 * f1 + rnorm(n, sd = .50),
  B1 = .86 * f2 + rnorm(n, sd = .48),
  B2 = .83 * f2 + rnorm(n, sd = .51),
  B3 = .84 * f2 + rnorm(n, sd = .50)
)

overlap_model <- '
  F1 =~ A1 + A2 + A3
  F2 =~ B1 + B2 + B3
'

overlap_cfa <- nomo_cfa(overlap_model, data = overlap)
overlap_rel <- nomo_reliability(overlap_cfa)
overlap_val <- nomo_validity(overlap_cfa, htmt = "both")
```

```{r overlap-summary}
summary(overlap_rel)
summary(overlap_val)
```

Here, strong omega and AVE do not erase a high latent correlation or HTMT2 value.
The package therefore recommends investigating theoretical distinctiveness, item
content, cross-construct overlap, and the intended construct boundary. It does not
automatically merge factors or delete indicators.

```{r overlap-plot}
plot(overlap_val, type = "discriminant")
```

# Ordered indicators: keep the score scale explicit

For ordered Likert-type indicators, `nomo_cfa()` uses an ordered-data CFA workflow
when the indicators are declared ordered. `nomo_reliability()` then distinguishes the
observed ordinal-score reliability estimand from the hypothetical latent-response
scale.

```{r ordinal-example}
set.seed(2029)
n <- 500
f <- rnorm(n)
z <- data.frame(
  I1 = .82 * f + rnorm(n, sd = .60),
  I2 = .79 * f + rnorm(n, sd = .62),
  I3 = .76 * f + rnorm(n, sd = .65),
  I4 = .80 * f + rnorm(n, sd = .60)
)

ordinal <- as.data.frame(lapply(z, function(x) {
  ordered(cut(x, breaks = c(-Inf, -.8, -.25, .25, .8, Inf), labels = 1:5))
}))

ordinal_cfa <- nomo_cfa(
  'F =~ I1 + I2 + I3 + I4',
  data = ordinal,
  ordered = names(ordinal)
)
ordinal_rel <- nomo_reliability(
  ordinal_cfa,
  ordinal_scale = TRUE,
  include_alpha = TRUE
)
summary(ordinal_rel)
```

Observed-scale omega is available for the practical ordinal composite. Observed-scale
alpha is deliberately reported as unavailable in this ordered-CFA estimand rather than
silently replacing it with a different definition. The point is not to maximize the
number of coefficients reported; it is to keep the score and estimand clear.

# Reading the evidence as an argument

A useful measurement conclusion is rarely "all cutoffs passed." A stronger conclusion
identifies what each analysis contributes and what remains uncertain. The workflow is
therefore designed around four questions:

- **Structure:** Is the hypothesized factor model a reasonable representation of the data?
- **Precision:** Is the intended score sufficiently reliable for its use, conditional on that model?
- **Convergence:** Do indicators substantially represent the intended construct?
- **Separation:** Are constructs that theory distinguishes also distinguishable empirically?

The resulting decision logs preserve observations, references, recommendations, and
researcher decisions without automatically changing the scale or model.

# Research basis

The reliability workflow follows the literature recommending model-based omega for
congeneric measurement and cautioning against mechanical use of coefficient alpha,
including Dunn, Baguley, and Brunsden (2014), Flora (2020), and Bell, Chalmers, and
Flora (2024). Ordered-score reliability follows the distinction emphasized by Green
and Yang (2009) and implemented in `semTools`.

The validity workflow treats AVE as convergent evidence rather than reliability,
following Fornell and Larcker (1981). For construct separation, HTMT-family evidence
is prioritized because simulation research shows that older criteria can miss overlap
(Henseler, Ringle, & Sarstedt, 2015). HTMT2 is emphasized for congeneric measurement
because it relaxes the original HTMT tau-equivalence assumption (Roemer, Schuberth, &
Henseler, 2021). Fornell-Larcker output remains available only as optional historical/
supporting information.
