---
title: "From CFA to a defensible measurement model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From CFA to a defensible measurement model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(nomologR)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.8
)
```

# Why measurement evidence is staged

A confirmatory factor model, a reliability coefficient, and a discriminant-validity
statistic answer different questions. Treating them as interchangeable checkboxes
can produce misleading conclusions. `nomologR` therefore keeps the workflow staged:

1. `nomo_cfa()` — **Does the proposed measurement structure reproduce the data reasonably, and where is there model strain?**
2. `nomo_reliability()` — **How consistently does the intended score represent its target under that measurement model?**
3. `nomo_validity()` — **Do indicators converge on their intended constructs, and are theoretically distinct constructs empirically distinguishable?**

The sequence is evidence-guided rather than pass/fail. A reference value can trigger
inspection, but it never automatically deletes an indicator, merges constructs, or
establishes construct validity.

# A healthy two-factor model

We begin with a known two-factor population in which both factors are well measured
and moderately correlated.

```{r healthy-data}
set.seed(2026)
n <- 500

f1 <- rnorm(n)
f2 <- .30 * f1 + sqrt(1 - .30^2) * rnorm(n)

healthy <- data.frame(
  A1 = .82 * f1 + rnorm(n, sd = .55),
  A2 = .79 * f1 + rnorm(n, sd = .58),
  A3 = .76 * f1 + rnorm(n, sd = .62),
  A4 = .80 * f1 + rnorm(n, sd = .57),
  B1 = .83 * f2 + rnorm(n, sd = .54),
  B2 = .78 * f2 + rnorm(n, sd = .60),
  B3 = .75 * f2 + rnorm(n, sd = .63),
  B4 = .81 * f2 + rnorm(n, sd = .56)
)

model <- '
  F1 =~ A1 + A2 + A3 + A4
  F2 =~ B1 + B2 + B3 + B4
'
```

```{r healthy-fit}
cfa <- nomo_cfa(model, data = healthy)

# Modest resampling here keeps vignette build time reasonable.
# Use more resamples for final reporting.
rel <- nomo_reliability(
  cfa,
  ci = "bootstrap",
  ci_boot = 100,
  ci_seed = 2026
)
val <- nomo_validity(cfa, htmt = "both")
```

The compact summaries are intentionally different from the raw engine output. They
surface the evidence most relevant to the user's next decision without repeating the
same full loading table in several places.

```{r healthy-tables}
summary(rel)
summary(val)
```

The figures also avoid duplication. Item-level loading evidence belongs to the CFA;
reliability gets a coefficient plot; convergent validity gets an AVE plot; and
construct separation gets an HTMT-family plot.

```{r healthy-loading-plot}
plot(cfa, type = "loadings")
```

```{r healthy-reliability-plot}
plot(rel)
```

```{r healthy-ave-plot}
plot(val, type = "ave")
```

```{r healthy-discriminant-plot}
plot(val, type = "discriminant")
```

A healthy result contributes several complementary pieces of evidence. It does not
turn the measurement model into a permanently "validated" object. Validity remains
an accumulating argument tied to construct interpretation, population, setting, and
intended use.

# Good global fit does not guarantee good measurement

This is a critical teaching case. The following one-factor model is correctly
specified, but its indicators are weak measures of the latent variable.

```{r weak-data}
set.seed(2027)
n <- 650
f <- rnorm(n)

weak <- data.frame(
  W1 = .30 * f + rnorm(n, sd = .95),
  W2 = .34 * f + rnorm(n, sd = .94),
  W3 = .28 * f + rnorm(n, sd = .96),
  W4 = .32 * f + rnorm(n, sd = .95),
  W5 = .31 * f + rnorm(n, sd = .95)
)

weak_cfa <- nomo_cfa(
  'Weak =~ W1 + W2 + W3 + W4 + W5',
  data = weak
)
weak_rel <- nomo_reliability(weak_cfa)
weak_val <- nomo_validity(weak_cfa, htmt = "none")
```

```{r weak-fit}
weak_cfa$fit_evidence
```

Because the one-factor structure is the correct covariance model, global fit can look
quite good. That does **not** imply that the items strongly represent the construct or
that the resulting score is precise.

```{r weak-measurement}
summary(weak_rel)
summary(weak_val)
```

This is why `nomologR` does not collapse measurement quality into a single global-fit
verdict. Weak standardized loadings, low AVE, and weak model-based reliability remain
important even when CFI, RMSEA, or SRMR look favorable.

# Strong reliability does not guarantee construct separation

The reverse problem is also possible. Two constructs can each be measured very
consistently while remaining difficult to distinguish empirically.

```{r overlap-data}
set.seed(2028)
n <- 700
f1 <- rnorm(n)
f2 <- .94 * f1 + sqrt(1 - .94^2) * rnorm(n)

overlap <- data.frame(
  A1 = .86 * f1 + rnorm(n, sd = .48),
  A2 = .83 * f1 + rnorm(n, sd = .51),
  A3 = .84 * f1 + rnorm(n, sd = .50),
  B1 = .86 * f2 + rnorm(n, sd = .48),
  B2 = .83 * f2 + rnorm(n, sd = .51),
  B3 = .84 * f2 + rnorm(n, sd = .50)
)

overlap_model <- '
  F1 =~ A1 + A2 + A3
  F2 =~ B1 + B2 + B3
'

overlap_cfa <- nomo_cfa(overlap_model, data = overlap)
overlap_rel <- nomo_reliability(overlap_cfa)
overlap_val <- nomo_validity(overlap_cfa, htmt = "both")
```

```{r overlap-summary}
summary(overlap_rel)
summary(overlap_val)
```

Here, strong omega and AVE do not erase a high latent correlation or HTMT2 value.
The package therefore recommends investigating theoretical distinctiveness, item
content, cross-construct overlap, and the intended construct boundary. It does not
automatically merge factors or delete indicators.

```{r overlap-plot}
plot(overlap_val, type = "discriminant")
```

# Ordered indicators: keep the score scale explicit

For ordered Likert-type indicators, `nomo_cfa()` uses an ordered-data CFA workflow
when the indicators are declared ordered. `nomo_reliability()` then distinguishes the
observed ordinal-score reliability estimand from the hypothetical latent-response
scale.

```{r ordinal-example}
set.seed(2029)
n <- 500
f <- rnorm(n)
z <- data.frame(
  I1 = .82 * f + rnorm(n, sd = .60),
  I2 = .79 * f + rnorm(n, sd = .62),
  I3 = .76 * f + rnorm(n, sd = .65),
  I4 = .80 * f + rnorm(n, sd = .60)
)

ordinal <- as.data.frame(lapply(z, function(x) {
  ordered(cut(x, breaks = c(-Inf, -.8, -.25, .25, .8, Inf), labels = 1:5))
}))

ordinal_cfa <- nomo_cfa(
  'F =~ I1 + I2 + I3 + I4',
  data = ordinal,
  ordered = names(ordinal)
)
ordinal_rel <- nomo_reliability(
  ordinal_cfa,
  ordinal_scale = TRUE,
  include_alpha = TRUE
)
summary(ordinal_rel)
```

Observed-scale omega is available for the practical ordinal composite. Observed-scale
alpha is deliberately reported as unavailable in this ordered-CFA estimand rather than
silently replacing it with a different definition. The point is not to maximize the
number of coefficients reported; it is to keep the score and estimand clear.

# Continuing the exploratory example

The [exploratory walkthrough](exploratory-workflow.html) flagged a cross-loading
item (`a5`) and a weak item (`b5`) in `nomo_demo_continuous`. Suppose the
researcher prespecifies the intended two-factor simple structure with all ten
items and fits it as a CFA:

```{r demo-cfa}
demo_model <- nomo_model(list(
  A = c("a1", "a2", "a3", "a4", "a5"),
  B = c("b1", "b2", "b3", "b4", "b5")
))

demo_cfa <- nomo_cfa(demo_model, data = nomo_demo_continuous)
demo_cfa$fit_evidence
head(demo_cfa$top_modification_indices, 5)
```

Three lessons carry over from the exploratory stage:

- **Modification indices are quarantined.** The largest index points toward the
  cross-loading that the data-generating model contains, but `nomologR` does not
  add it. Allowing a cross-loading, revising the item, or evaluating a model
  without it is a theoretical decision; freeing parameters because an index is
  large is capitalization on chance (MacCallum, Roznowski, & Necowitz, 1992).
- **The same sample is not independent confirmation.** This CFA uses the data
  that suggested the structure. `nomo_split()` or a new sample provides a
  stronger test.
- **Missing data are handled explicitly.** By default lavaan uses complete cases
  here; `missing = "fiml"` requests full-information maximum likelihood for
  continuous indicators, and the decision is recorded.

# Comparing models with a recorded rationale

`nomo_compare()` evaluates competing measurement models side by side. It
requires a rationale, checks whether the models are nested, reports the
difference test that matches the estimator when they are, and never selects a
model automatically.

## A cross-loading suggested by the results

The largest modification index above pointed to a loading of `a5` on factor B.
Because that idea came from the results rather than from theory stated in
advance, the comparison is labeled post hoc:

```{r compare-cross}
cross_cfa <- nomo_cfa(
  "A =~ a1 + a2 + a3 + a4 + a5\nB =~ b1 + b2 + b3 + b4 + b5 + a5",
  data = nomo_demo_continuous
)

cross_comparison <- nomo_compare(
  simple_structure = demo_cfa,
  cross_loading = cross_cfa,
  rationale = "The largest modification index suggested that a5 also reflects factor B.",
  origin = "post_hoc"
)
cross_comparison
```

```{r compare-values, include=FALSE}
fmt_p <- function(p) {
  if (!is.finite(p)) return("unavailable")
  if (p < .001) "< .001" else paste("=", formatC(p, format = "f", digits = 3))
}
cc <- cross_comparison$comparisons
```

Estimating the cross-loading reduces misfit (chi-square difference =
`r sprintf("%.1f", cc$chisq_diff)`, df = `r cc$df_diff`, p `r fmt_p(cc$p_value)`;
CFI changes by `r sprintf("%+.3f", cc$delta_cfi)`). That is exactly the feature
built into these simulated data, but with real data the evidence alone does not
settle the question. Does the wording of `a5` plausibly reflect both
constructs? Because the comparison is post hoc, the decision log records it as
such and recommends confirming the retained model in independent data.

## Is a weak item needed?

`b5` loads weakly on B. Dropping `b5` from the model changes the data being
modeled, so a model without the column cannot be tested against the full model.
Instead, keep `b5` and fix its loading to zero:

```{r compare-b5}
b5_zero_cfa <- nomo_cfa(
  "A =~ a1 + a2 + a3 + a4 + a5\nB =~ b1 + b2 + b3 + b4 + 0*b5",
  data = nomo_demo_continuous
)

b5_comparison <- nomo_compare(
  full = demo_cfa,
  b5_loading_zero = b5_zero_cfa,
  rationale = paste(
    "Evaluate whether the weakly loading item b5 contributes to factor B",
    "before deciding whether to keep it."
  )
)
summary(b5_comparison)
```

```{r b5-values, include=FALSE}
bc <- b5_comparison$comparisons
b5_loading <- b5_comparison$loadings$full[b5_comparison$loadings$item == "b5"]
```

Fixing the loading to zero worsens fit (chi-square difference =
`r sprintf("%.1f", bc$chisq_diff)`, df = `r bc$df_diff`, p `r fmt_p(bc$p_value)`),
so `b5` is statistically related to B. Its standardized loading, however, is
only `r sprintf("%.2f", b5_loading)`. With `r b5_comparison$models$n_used[[1]]`
cases, even a weak relation is detectable; statistical detectability is not the
same as an adequate indicator. The decision still depends on whether `b5` covers
content the construct needs.

```{r compare-loadings-plot}
plot(b5_comparison, type = "loadings")
```

The summary's measurement evidence carries a note: omega for B in the
zero-loading model still includes `b5` in the composite, so it does not
describe a shortened four-item scale. To see that scale's reliability, fit the
model without `b5`. Because the two models contain different observed
variables, `nomo_compare()` reports this comparison descriptively, without a
difference test or information criteria:

```{r compare-shortened}
shortened_cfa <- nomo_cfa(
  "A =~ a1 + a2 + a3 + a4 + a5\nB =~ b1 + b2 + b3 + b4",
  data = nomo_demo_continuous
)

shortened <- nomo_compare(
  full = demo_cfa,
  four_item_b = shortened_cfa,
  rationale = "Describe the reliability of a four-item B scale alongside the full scale."
)

nomo_table(shortened, "comparisons")[, c("model", "relation", "test_available", "ic_available")]

omega_rows <- nomo_table(shortened, "evidence")
omega_rows[omega_rows$metric == "omega", c("model", "construct", "estimate")]
```

Together these results separate three questions that are easy to blur: whether
an item is statistically related to its factor, whether it is a strong
indicator, and how the score's reliability changes without it. None of them
alone decides whether the item stays.

# Reading the evidence as an argument

A useful measurement conclusion is rarely "all cutoffs passed." A stronger conclusion
identifies what each analysis contributes and what remains uncertain. The workflow is
therefore designed around four questions:

- **Structure:** Is the hypothesized factor model a reasonable representation of the data?
- **Precision:** Is the intended score sufficiently reliable for its use, conditional on that model?
- **Convergence:** Do indicators substantially represent the intended construct?
- **Separation:** Are constructs that theory distinguishes also distinguishable empirically?

The resulting decision logs preserve observations, references, recommendations, and
researcher decisions without automatically changing the scale or model.

# Research basis

The reliability workflow follows the literature recommending model-based omega for
congeneric measurement and cautioning against mechanical use of coefficient alpha,
including Dunn, Baguley, and Brunsden (2014), Flora (2020), and Bell, Chalmers, and
Flora (2024). Ordered-score reliability follows the distinction emphasized by Green
and Yang (2009) and implemented in `semTools`.

The validity workflow treats AVE as convergent evidence rather than reliability,
following Fornell and Larcker (1981). For construct separation, HTMT-family evidence
is prioritized because simulation research shows that older criteria can miss overlap
(Henseler, Ringle, & Sarstedt, 2015). HTMT2 is emphasized for congeneric measurement
because it relaxes the original HTMT tau-equivalence assumption (Roemer, Schuberth, &
Henseler, 2021). Fornell-Larcker output remains available only as optional historical/
supporting information.
