---
title: "Get started with nomologR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with nomologR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

```{r library}
library(nomologR)
```

# Who nomologR is for

`nomologR` is written for three audiences:

- **graduate students** (master's and doctoral) who are learning scale
  development and construct validation and need to understand *why* each
  analysis is performed, not only how to run it;
- **faculty and advisors** who teach measurement and want output that explains
  its own reasoning; and
- **researchers** who need a defensible, reproducible record of how a measure
  was evaluated.

The package coordinates established engines — `psych`, `EFAtools`, `lavaan`,
and `semTools` — and adds a teaching layer: diagnostics framed as evidence,
explanations tied to the methodological literature, and a record of every
consequential researcher decision.

# The question behind every stage

At each step, `nomologR` tries to help answer one question:

> **What does this result tell me about the claim that these observations
> measure the construct I say they measure, and what should I investigate
> next?**

The package flags, explains, and documents. It never silently deletes items,
respecifies models, or declares a scale "valid."

# What nomologR assumes about the items

Every analysis in `nomologR` assumes *reflective* measurement: the construct
causes the responses, so its items are interchangeable manifestations of it and
should correlate with one another. That assumption is what makes factor
loadings, internal consistency, AVE, HTMT, and measurement invariance
meaningful, and what makes a weakly loading item worth reviewing (Bollen &
Lennox, 1991).

Some measures are built the other way. In a *formative* measure, the indicators
define the construct, as income, education, and occupation define
socioeconomic status. They need not correlate, and dropping one changes what
is measured (Diamantopoulos & Winklhofer, 2001). The tools here do not apply
to such a measure, and applying them misleads: low internal consistency is
expected, and deleting a "weak" indicator removes part of the construct.
Decide which model a measure follows before choosing tools; Jarvis,
MacKenzie, and Podsakoff (2003) and MacKenzie, Podsakoff, and Jarvis (2005)
give criteria. Whether formative latent variables are meaningful at all is
still debated (Howell, Breivik, & Wilcox, 2007; Edwards, 2011; Bollen &
Diamantopoulos, 2017), and `nomologR` does not fit them. A formative index can
still enter a nomological network as an observed variable in `nomo_network()`,
and its content can still be reviewed with `contentvalidR`.

# The workflow at a glance

| Stage | Question it answers | Functions | Walkthrough |
|---|---|---|---|
| 1. Item and data audit | What do the items and responses look like? | `nomo_screen()` | [From item audit to exploratory structure](exploratory-workflow.html) |
| 2. Dimensionality | How many latent dimensions deserve investigation? | `nomo_factors()` | [From item audit to exploratory structure](exploratory-workflow.html) |
| 3. Exploratory structure | What does a requested factor solution look like? | `nomo_efa()` | [From item audit to exploratory structure](exploratory-workflow.html) |
| 4. Confirmatory measurement model | Does a prespecified model reproduce the data, and where is there strain? | `nomo_model()`, `nomo_split()`, `nomo_cfa()` | [From CFA to a defensible measurement model](measurement-model-evidence.html) |
| 5. Reliability and construct-validity evidence | How precise are scores, and are constructs distinguishable? | `nomo_reliability()`, `nomo_validity()` | [From CFA to a defensible measurement model](measurement-model-evidence.html) |
| 6. Generalizability | Is the construct measured comparably across groups and occasions? | `nomo_invariance()`, `nomo_invariance_longitudinal()`, `nomo_partial()` | [Measurement invariance](measurement-invariance.html) |
| 7. Nomological network | Does the construct relate to others as theory predicted? | `nomo_hypotheses()`, `nomo_network()` | [Theory-specified nomological networks](nomological-network.html) |
| Guided workflow | How do the stages fit together with explicit decisions? | `nomo_run()` | [Guided workflow](guided-workflow.html) |
| Starting from content review | What does empirical evidence add to an expert review of the items? | `nomo_screen()` and `nomo_run()` with a `contentvalidR` handoff | [From content review to empirical screening](content-review.html) |
| Reporting | How do I archive the evidence and decisions? | `nomo_report()`, `nomo_table()` | [Archiving a workflow](reproducible-report.html) |

# Teaching datasets

The examples and walkthroughs use simulated datasets whose population models
are documented. Because the truth is known, you can check what the package
reports against what actually generated the data — something real data never
allow.

| Dataset | Design | What it teaches |
|---|---|---|
| `nomo_demo_continuous` | Two correlated factors, five candidate items each, 500 cases | A cross-loading item (`a5`), a weak item (`b5`), and a small amount of missing data |
| `nomo_demo_ordinal` | The same latent responses cut into five ordered categories | Polychoric correlations, WLSMV estimation, and ordinal reliability |
| `nomo_demo_network` | Three constructs, an observed outcome, two administration groups, 800 cases | Theory-specified predictions, equivalence regions, replication, and a known source of scalar non-invariance |
| `nomo_demo_walkthrough` | Twelve items an expert panel reviewed in `contentvalidR`, two facets, two cohorts, 400 cases | Items that pass content review and fail empirically, and the reverse; reverse-worded items; a cohort difference |

See `?nomo_demo_continuous`, `?nomo_demo_ordinal`, `?nomo_demo_network`, and
`?nomo_demo_walkthrough` for the full population models.

# A first look

An item audit describes the data without changing it:

```{r first-screen}
scr <- nomo_screen(nomo_demo_continuous)
scr
```

Factor-retention evidence triangulates several criteria rather than trusting a
single rule:

```{r first-factors}
fac <- nomo_factors(nomo_demo_continuous, seed = 2026)
fac
```

The population model for these data has two factors, and the retention evidence
points to investigating two. Notice the wording: the evidence supports
*investigating* a two-factor solution; it does not prove that the scale "has"
two factors.

# How to read nomologR output

- **Observation, reason, options, consequence.** Recommendations state what was
  found, why it may matter, what defensible choices exist, and what each choice
  changes downstream.
- **Reference values are prompts, not verdicts.** Familiar numbers such as a
  loading of .40 or a CFI of .95 flag evidence for review. Items receive
  `KEEP`, `REVIEW`, or `STRONG REVIEW`, never `DELETE`. The references live in
  `nomo_defaults()`, and changing them is a visible researcher choice.
- **One flag wording on screen.** Printed output, plots, and reports use one
  flag wording: "review", then "concern", and no flag when no reference fired
  (a plot legend calls that "none"). The returned tables keep each analysis's
  own values (`KEEP`, `REVIEW`, and `STRONG REVIEW` for loadings; `info`,
  `review`, and `concern` for evidence), so code that filters on them is
  unaffected.
- **Historical methods are labeled.** Techniques you will meet in published
  work — eigenvalues greater than one, coefficient alpha, the Fornell–Larcker
  comparison — are shown as context where useful, qualified, and kept separate
  from contemporary evidence.
- **Engine results are never hidden.** Every result keeps the underlying
  `psych` or `lavaan` object (usually in `$fit`) for direct inspection.
- **Decisions are recorded.** Each result carries a decision log, and
  `nomo_run()` records the researcher's decisions and rationales across stages.

# A suggested learning path

1. [From item audit to exploratory structure](exploratory-workflow.html) —
   screening, factor retention, and EFA.
2. [From CFA to a defensible measurement model](measurement-model-evidence.html)
   — confirmatory fit, reliability, and convergent and discriminant evidence.
3. [Measurement invariance](measurement-invariance.html) — comparability across
   groups, localized strain, and researcher-controlled partial invariance.
4. [Theory-specified nomological networks](nomological-network.html) —
   predictions, equivalence regions, and replication.
5. [Guided workflow](guided-workflow.html) — the stages combined with explicit
   decisions.
6. [From content review to empirical screening](content-review.html) — the
   items a `contentvalidR` review carried forward, screened, including where
   the two stages disagree.
7. [Archiving a workflow](reproducible-report.html) — a reproducible report.

Throughout, the [research basis](research-basis.html) article explains where
each method came from, what the contemporary literature recommends, and what
`nomologR` implements. Every function's help page lists its references.

# Where nomologR is heading

The next release, `v1.0.0`, is planned together with
[`contentvalidR`](https://github.com/JUhalt/contentvalidR), which covers the
content-review stage before this package's. At 1.0.0 both packages' interfaces
become stable, and the handoff between them becomes a supported contract.
Larger extensions, such as IRT, multiple imputation, and formative models,
are candidates for later 1.x releases. See the
[roadmap](https://github.com/JUhalt/nomologR/blob/master/ROADMAP.md) and the
[v1.0.0 milestone](https://github.com/JUhalt/nomologR/milestone/5).

# Citing nomologR

```{r citation}
citation("nomologR")
```
