---
title: "Theory-Specified Nomological Networks with nomologR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Theory-Specified Nomological Networks with nomologR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(nomologR)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

# Why theory comes before the network

`nomologR` treats nomological evidence as a test of **explicit theoretical
predictions**, not as a search for statistically significant relationships.

The intended sequence is:

1. establish a defensible measurement model;
2. specify the expected relations before inspecting the structural results;
3. fit the full model;
4. compare estimates and uncertainty with the prespecified theoretical region;
5. distinguish theory strain, imprecision, measurement problems, and post-hoc
   exploration.

# The teaching study

`nomo_demo_network` simulates a validation study of an **Agency** measure. Two
related constructs were also measured — **Persistence** and **Social
desirability** — along with an observed **Performance** outcome.

The population model is known (see `?nomo_demo_network`):

- Agency predicts Persistence (standardized coefficient .45);
- Agency predicts Performance (.40);
- Persistence has **no** direct effect on Performance once Agency is accounted
  for (0);
- Agency and Social desirability are unrelated (0).

In real research you would not know these values. They are shown here so you
can see how `nomologR` evidence behaves when a prediction is right, when it is
wrong, and when it is about a negligible relation.

# Specify theory before looking at results

```{r hypotheses}
h <- nomo_hypotheses(
  "Agency -> Persistence" = positive(min = .20),
  "Agency <-> SocialDesirability" = negligible(within = c(-.15, .15)),
  "Agency -> Performance" = positive(),
  "Persistence -> Performance" = positive(min = .20)
)

h
```

`A -> B` means a directed structural path. `A <-> B` means an association
without asserting causal direction.

The fourth prediction is plausible-sounding — persistent people perform better —
but it is **not** true in the population model. It is included on purpose.

The negligible-effect region of ±.15 for Social desirability is a
**researcher-specified smallest effect size of interest (SESOI)**. It must be
justified by theory or study design before the analysis; `nomologR` never
invents one. A bare `negligible()` records a theoretical expectation, but
`nomologR` will refuse to treat `p > .05` as confirmation of negligibility.

# Fit the latent network

```{r fit}
model <- nomo_model(list(
  Agency = c("ag1", "ag2", "ag3", "ag4"),
  Persistence = c("pe1", "pe2", "pe3", "pe4"),
  SocialDesirability = c("sd1", "sd2", "sd3")
))

net <- nomo_network(
  model,
  data = nomo_demo_network,
  hypotheses = h
)

net
nomo_table(net, "fit")
```

The measurement model supplies only the factor definitions. Theory-specified
paths absent from that model are added transparently, and the exact fitted
syntax and unchanged `lavaan` fit remain available in the returned object.

# Read relation-level evidence

```{r evidence}
evidence <- nomo_table(net, "hypotheses")
evidence[, c(
  "id", "relation", "theoretical_region", "estimate",
  "ci_lower", "ci_upper", "equivalence_supported", "concordance"
)]
```

```{r evidence-values, include=FALSE}
ev <- function(id, col) evidence[[col]][evidence$id == id]
fmt <- function(x) sprintf("%.2f", x)
```

Reading each row:

- **H1** (`Agency -> Persistence`): estimate `r fmt(ev("H1", "estimate"))`,
  interval [`r fmt(ev("H1", "ci_lower"))`, `r fmt(ev("H1", "ci_upper"))`],
  classified **`r ev("H1", "concordance")`** with the prediction of at least
  .20.
- **H2** (`Agency <-> SocialDesirability`): estimate
  `r fmt(ev("H2", "estimate"))`. The 90% equivalence interval
  [`r fmt(ev("H2", "equivalence_ci_lower"))`,
  `r fmt(ev("H2", "equivalence_ci_upper"))`] lies inside the ±.15 region, so
  the negligible prediction is supported by equivalence evidence — not merely
  by a non-significant p-value.
- **H3** (`Agency -> Performance`): estimate `r fmt(ev("H3", "estimate"))`,
  classified **`r ev("H3", "concordance")`**.
- **H4** (`Persistence -> Performance`): estimate `r fmt(ev("H4", "estimate"))`,
  interval [`r fmt(ev("H4", "ci_lower"))`, `r fmt(ev("H4", "ci_upper"))`],
  classified **`r ev("H4", "concordance")`** with the predicted region.

Every relation also retains its standard error, p-value, evidence scope, a-priori
versus post-hoc provenance, and a measurement-context flag. The goal is not to
produce a single "validity score."

# A marginal association is not a structural prediction

Why might a researcher have expected H4? Look at the simple correlation between
a Persistence composite and Performance:

```{r marginal}
persistence_mean <- rowMeans(nomo_demo_network[, c("pe1", "pe2", "pe3", "pe4")])
cor(persistence_mean, nomo_demo_network$Performance)
```

The marginal correlation is positive, because both Persistence and Performance
depend on Agency. The theory-specified path asks a different question: does
Persistence predict Performance *given* Agency? It does not. Treating a table
of observed correlations as the nomological network would have produced the
wrong conclusion.

# Negligible predictions and equivalence evidence

For:

```{r negligible-example}
negligible(within = c(-.15, .15))
```

the researcher has supplied a quantitative negligible-effect region. With the
default `equivalence_alpha = .05`, `nomo_network()` evaluates a 90% equivalence
confidence interval, corresponding to the two one-sided tests procedure. This
is intentionally different from declaring a relation negligible because an
ordinary null-hypothesis test was not significant.

# Criterion and predictive evidence

`Performance` is an observed variable, while Agency is latent. The evidence table
labels such relations explicitly:

```{r scope}
evidence[, c("id", "relation", "evidence_scope")]
```

Observed outcomes keep their observed-variable estimand; `nomologR` does not
silently insert measurement-error corrections.

# Measurement context

```{r measurement}
nomo_table(net, "measurement")
```

A relation can look inconsistent with theory because the estimated relationship
truly differs from the prediction, or because weak measurement, improper
solutions, poor fit, or large uncertainty limit what the structural result can
say. The measurement context stays beside the theory evidence so the two are not
confused.

# Calibration and validation

An internal split can be created explicitly:

```{r split}
s <- nomo_split(nomo_demo_network, validation_prop = .40, seed = 2026)

net_rep <- nomo_network(
  model,
  data = s,
  hypotheses = h
)

replication <- nomo_table(net_rep, "replication")
replication[, c(
  "id", "relation", "primary_estimate", "validation_estimate",
  "validation_concordance", "replication_status"
)]
```

```{r replication-values, include=FALSE}
rs <- function(id) replication$replication_status[replication$id == id]

h4_reading <- switch(
  rs("H4"),
  sign_change_within_uncertainty = paste(
    "neither sample's confidence interval excludes zero, so the sign change is",
    "attributed to sampling uncertainty rather than to a substantive discrepancy"
  ),
  direction_not_replicated = paste(
    "only one sample's confidence interval excludes zero, so the direction was",
    "not replicated, but no reversal is established"
  ),
  sign_reversal = paste(
    "both confidence intervals exclude zero on opposite sides, which is what a",
    "reversal requires"
  ),
  "the status is read from both confidence intervals, not the point estimates alone"
)
```

The calibration and validation subsets receive the **same prespecified fitted
model**; `nomologR` does not respecify the validation model to rescue a primary
result. Reading the statuses:

- H1 and H3 are **`r rs("H1")`** and **`r rs("H3")`**.
- H2 is **`r rs("H2")`**: with only `r nrow(s$validation)` validation cases, the
  equivalence interval is too wide to confirm negligibility. Splitting a sample
  buys independence at the cost of precision.
- H4 is **`r rs("H4")`**. The population path is zero, and the two point
  estimates happen to differ in sign. Here `r h4_reading`.

A change in sign between samples needs care. Estimates scattered around a null
relation differ in sign about half the time, so `nomologR` does not call a sign
change a **reversal** from the point estimates. It reads the 95 percent
confidence intervals:

| Status | Intervals | Meaning |
|---|---|---|
| `sign_reversal` | Both exclude zero, on opposite sides | Each sample on its own supports a different direction |
| `direction_not_replicated` | Exactly one excludes zero | The other sample does not support that direction, nor establish the opposite |
| `sign_change_within_uncertainty` | Neither excludes zero | Neither sample distinguishes the relation from zero |

None of these turns a non-significant path into evidence of *no* relation. That
claim needs a `negligible(within = ...)` prediction with an equivalence region,
the kind of prediction H2 makes (Lakens, Scheel, & Isager, 2018).

External validation data can instead be supplied with `validation_data =`, which
is generally stronger evidence than an internal split.

# Figures

```{r effects-plot}
plot(net, type = "effects")
```

```{r replication-plot}
plot(net_rep, type = "replication")
```

These are evidence summaries, not automated theory verdicts.

# Interpretation rule

`nomologR` keeps **measurement evidence, theory concordance, uncertainty,
replication, and researcher provenance visible at the same time**. A supported
prediction strengthens the validity argument for a particular interpretation of
scores; an unsupported one is information about the theory, the measure, or
both.

# Research basis

The nomological network as the core of construct validity comes from Cronbach
and Meehl (1955), and convergent and discriminant correlation patterns from
Campbell and Fiske (1959). Contemporary practice separates the measurement model
from structural relations (Anderson & Gerbing, 1988), treats validity as an
argument about score interpretation (Messick, 1995), evaluates negligible
predictions with equivalence procedures (Schuirmann, 1987; Lakens, Scheel, &
Isager, 2018), and distinguishes prespecified from post-hoc predictions (Nosek
et al., 2018). Full references are in `?nomo_network` and the
[research basis](research-basis.html) article.
