---
title: "Theory-Specified Nomological Networks with nomologR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Theory-Specified Nomological Networks with nomologR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

# Why theory comes before the network

`nomologR` treats nomological evidence as a test of **explicit theoretical
predictions**, not as a search for statistically significant relationships.

The intended sequence is:

1. establish a defensible measurement model;
2. specify the expected relations before inspecting the structural results;
3. fit the full model;
4. compare estimates and uncertainty with the prespecified theoretical region;
5. distinguish theory strain, imprecision, measurement problems, and post-hoc
   exploration.

# Specify theory

```{r}
h <- nomo_hypotheses(
  "Agency -> Persistence" = positive(min = .20),
  "Agency <-> SocialDesirability" =
    negligible(within = c(-.10, .10)),
  "Agency -> Performance" = positive()
)

h
```

`A -> B` means a directed structural path. `A <-> B` means an association
without asserting causal direction.

A bare `negligible()` records a theoretical expectation but does **not** supply
a smallest effect size of interest (SESOI). `nomologR` will therefore refuse to
treat `p > .05` as confirmation of negligibility.

# Fit the latent network

```{r}
model <- "
  Agency =~ a1 + a2 + a3
  Persistence =~ p1 + p2 + p3
"

net <- nomo_network(
  model,
  data = dat,
  hypotheses = h
)

net
summary(net)
```

If a theory-specified path is absent from the supplied measurement model,
`nomo_network()` can add it transparently. The exact fitted syntax and unchanged
`lavaan` fit remain available in the returned object.

# Read relation-level evidence

```{r}
nomo_table(net, "hypotheses")
```

Every relation retains:

- predicted direction or region;
- estimate;
- standard error;
- confidence interval;
- standardized estimate;
- p-value where available;
- concordance classification;
- measurement-context flag;
- a-priori versus post-hoc provenance.

The goal is not to produce a single "validity score."

# Negligible predictions and equivalence evidence

For:

```{r}
negligible(within = c(-.10, .10))
```

the researcher has supplied a quantitative negligible-effect region. With the
default `equivalence_alpha = .05`, `nomo_network()` evaluates a 90% equivalence
confidence interval. This is intentionally different from declaring a relation
negligible because an ordinary null-hypothesis test was not significant.

# Criterion and predictive evidence

Observed outcomes can participate in the same network:

```{r}
h <- nomo_hypotheses(
  "Agency -> Performance" = positive()
)
```

When `Performance` is an observed variable and `Agency` is latent, the evidence
table labels the relation as a latent-to-observed outcome path.

# Calibration and validation

An internal split can be created explicitly:

```{r}
s <- nomo_split(dat, validation_prop = .40, seed = 2026)

net <- nomo_network(
  model,
  data = s,
  hypotheses = h
)
```

The calibration and validation subsets receive the **same prespecified fitted
model**. `nomologR` does not respecify the validation model to rescue a primary
sample result.

External validation data can instead be supplied with `validation_data =`.

# Figures and manuscript-ready tables

```{r}
plot(net, "effects")
plot(net, "concordance")
plot(net, "fit")
plot(net, "replication")

nomo_table(net, "hypotheses")
nomo_table(net, "measurement")
nomo_table(net, "replication")
```

These are evidence summaries, not automated theory verdicts.

# Interpretation rule

A relation can be inconsistent with theory because the estimated relationship
truly differs from the prediction. But weak measurement, improper solutions,
poor global fit, or large uncertainty can also limit what the structural result
can say.

`nomologR` therefore keeps **measurement evidence, theory concordance,
uncertainty, replication, and researcher provenance visible at the same time**.
