---
title: "Theory-Specified Nomological Networks with nomologR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Theory-Specified Nomological Networks with nomologR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(nomologR)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

# Why theory comes before the network

`nomologR` treats nomological evidence as a test of **explicit theoretical
predictions**, not as a search for statistically significant relationships.

The intended sequence is:

1. establish a defensible measurement model;
2. specify the expected relations before inspecting the structural results;
3. fit the full model;
4. compare estimates and uncertainty with the prespecified theoretical region;
5. distinguish theory strain, imprecision, measurement problems, and post hoc
   exploration.

# The teaching study

`nomo_demo_network` simulates a validation study of an **Agency** measure. Two
related constructs were also measured — **Persistence** and **Social
desirability** — along with an observed **Performance** outcome.

The population model is known (see `?nomo_demo_network`):

- Agency predicts Persistence (standardized coefficient 0.45);
- Agency predicts Performance (0.40);
- Persistence has **no** direct effect on Performance once Agency is accounted
  for (0);
- Agency and Social desirability are unrelated (0).

In real research you would not know these values. They are shown here so you
can see how `nomologR` evidence behaves when a prediction is right, when it is
wrong, and when it is about a negligible relation.

# Specify theory before looking at results

```{r hypotheses}
h <- nomo_hypotheses(
  "Agency -> Persistence" = positive(min = .20),
  "Agency <-> SocialDesirability" = negligible(within = c(-.15, .15)),
  "Agency -> Performance" = positive(),
  "Persistence -> Performance" = positive(min = .20)
)

h
```

`A -> B` means a directed structural path. `A <-> B` means an association
without asserting causal direction.

The fourth prediction is plausible-sounding — persistent people perform better —
but it is **not** true in the population model. It is included on purpose.

The negligible-effect region of ±.15 for Social desirability is a
**researcher-specified smallest effect size of interest (SESOI)**. It must be
justified by theory or study design before the analysis; `nomologR` never
invents one. A bare `negligible()` records a theoretical expectation, but
`nomologR` will refuse to treat `p > .05` as confirmation of negligibility.

# Fit the latent network

```{r fit}
model <- nomo_model(list(
  Agency = c("ag1", "ag2", "ag3", "ag4"),
  Persistence = c("pe1", "pe2", "pe3", "pe4"),
  SocialDesirability = c("sd1", "sd2", "sd3")
))

net <- nomo_network(
  model,
  data = nomo_demo_network,
  hypotheses = h
)

net
nomo_table(net, "fit")
```

The measurement model supplies only the factor definitions. Theory-specified
paths absent from that model are added transparently, and the exact fitted
syntax and unchanged `lavaan` fit remain available in the returned object.

# Read relation-level evidence

```{r evidence}
evidence <- nomo_table(net, "hypotheses")
evidence[, c("id", "relation", "theoretical_region", "concordance")]
evidence[, c(
  "id", "estimate", "ci_lower", "ci_upper",
  "equivalence_ci_lower", "equivalence_ci_upper"
)]
```

```{r evidence-values, include=FALSE}
ev <- function(id, col) evidence[[col]][evidence$id == id]
# Standardized estimates in the prose: two decimals, rounded half away from
# zero as the package's printed output is, and never a negative zero. A path
# coefficient keeps its leading zero, as a standardized loading does; a
# correlation, which cannot exceed 1, drops it.
fmt <- function(x, correlation = FALSE) {
  r <- sign(x) * floor(abs(x) * 100 + 0.5 + 1e-9) / 100
  out <- formatC(abs(r), format = "f", digits = 2)
  if (correlation) out <- sub("^0[.]", ".", out)
  paste0(if (r < 0) "-" else "", out)
}
```

The first table gives each prediction and how the evidence was classified, and
the second the estimates behind it. Reading each relation:

- **H1** (`Agency -> Persistence`): estimate `r fmt(ev("H1", "estimate"))`,
  interval [`r fmt(ev("H1", "ci_lower"))`, `r fmt(ev("H1", "ci_upper"))`],
  classified **`r ev("H1", "concordance")`** with the prediction of at least
  0.20.
- **H2** (`Agency <-> SocialDesirability`), a correlation: estimate
  `r fmt(ev("H2", "estimate"), correlation = TRUE)`. The 90% equivalence
  interval [`r fmt(ev("H2", "equivalence_ci_lower"), correlation = TRUE)`,
  `r fmt(ev("H2", "equivalence_ci_upper"), correlation = TRUE)`] lies inside
  the ±.15 region, so
  the negligible prediction is supported by equivalence evidence — not merely
  by a non-significant *p* value.
- **H3** (`Agency -> Performance`): estimate `r fmt(ev("H3", "estimate"))`,
  classified **`r ev("H3", "concordance")`**.
- **H4** (`Persistence -> Performance`): estimate `r fmt(ev("H4", "estimate"))`,
  interval [`r fmt(ev("H4", "ci_lower"))`, `r fmt(ev("H4", "ci_upper"))`],
  classified **`r ev("H4", "concordance")`** with the predicted region.

Every relation also retains its standard error, *p* value, evidence scope, a
priori versus post hoc provenance, and a measurement-context flag. The goal is
not to produce a single "validity score."

# A marginal association is not a structural prediction

Why might a researcher have expected H4? Look at the simple correlation between
a Persistence composite and Performance:

```{r marginal}
persistence_mean <- rowMeans(nomo_demo_network[, c("pe1", "pe2", "pe3", "pe4")])
cor(persistence_mean, nomo_demo_network$Performance)
```

The marginal correlation is positive, because both Persistence and Performance
depend on Agency. The theory-specified path asks a different question: does
Persistence predict Performance *given* Agency? It does not. Treating a table
of observed correlations as the nomological network would have produced the
wrong conclusion.

# Negligible predictions and equivalence evidence

For:

```{r negligible-example}
negligible(within = c(-.15, .15))
```

the researcher has supplied a quantitative negligible-effect region. With the
default `equivalence_alpha = .05`, `nomo_network()` evaluates a 90% equivalence
confidence interval, corresponding to the two one-sided tests procedure. This
is intentionally different from declaring a relation negligible because an
ordinary null-hypothesis test was not significant.

# Criterion and predictive evidence

`Performance` is an observed variable, while Agency is latent. The evidence table
labels such relations explicitly:

```{r scope}
evidence[, c("id", "relation", "evidence_scope")]
```

Observed outcomes keep their observed-variable estimand; `nomologR` does not
insert measurement-error corrections on its own. When a composite's reliability
is known, you can ask for one.

# A composite corrected for its unreliability

A scale mean carries its items' measurement error, and the relations it enters
are attenuated by it. A single-indicator latent variable corrects this. The
composite becomes the one indicator of a latent variable, and its error variance
is fixed at (1 − reliability) × its variance. The idea goes back to Spearman's
(1904) correction for attenuation. The SEM textbooks describe it (Hayduk, 1987;
Bollen, 1989), applied work in organizational research and marketing took it
up (Williams & Hazer, 1986), and Bagozzi and Heatherton (1994) called it the
"total aggregation" model. In samples of 30 to 200, Savalei (2019) found that
path analysis and single indicators whose reliability was fixed in advance, at
a value slightly above the true one, gave the most accurate estimates and the
most power; single indicators whose reliability was estimated from the same
data performed in between, and she recommended a fixed-reliability single
indicator in small samples. An omega from the same sample, as below,
corresponds to the data-estimated variant.

Here the Persistence mean replaces the Persistence factor. Its reliability is
omega from its own measurement model:

```{r single-indicator}
dat <- nomo_demo_network
dat$persistence <- rowMeans(dat[c("pe1", "pe2", "pe3", "pe4")])
rel <- nomo_reliability(nomo_cfa("Persistence =~ pe1 + pe2 + pe3 + pe4", dat))
h_si <- nomo_hypotheses("Agency -> persistence" = positive(min = .20))

as_observed <- nomo_network("Agency =~ ag1 + ag2 + ag3 + ag4", dat, h_si)
corrected <- nomo_network(
  "Agency =~ ag1 + ag2 + ag3 + ag4", dat, h_si,
  single_indicators = list(persistence = nomo_single_indicator(rel))
)
nomo_table(corrected, "single_indicators")[, c(
  "variable", "coefficient", "reliability", "variance", "error_variance"
)]
```

The population path is 0.45. The observed composite gives
`r fmt(as_observed$hypothesis_evidence$estimate)`, attenuated by its
unreliability. The single indicator gives
`r fmt(corrected$hypothesis_evidence$estimate)`, close to the latent model's
`r fmt(ev("H1", "estimate"))`.

A correction is only as good as the reliability behind it. So the network is
refitted with the reliability .05 and .10 lower and higher, and the log flags
any hypothesis whose evidence changes across that range:

```{r single-indicator-sensitivity}
nomo_table(corrected, "sensitivity")
```

Three choices matter:

- **Omega or alpha.** Coefficient alpha understates reliability when loadings
  differ, which overcorrects, so it is flagged for review.
- **Which errors the coefficient sees.** Any coefficient corrects only the error
  its design can detect. Internal consistency leaves transient error in, for
  example (DeShon, 1998).
- **Uncertainty in the reliability.** A reliability is itself an estimate.
  Standard errors that treat it as known are too small (Oberski & Satorra,
  2013). Give its standard error, or pass a bootstrapped `nomo_reliability()`
  result, and the intervals include that uncertainty.

# Measurement context

```{r measurement}
nomo_table(net, "measurement")
```

A relation can look inconsistent with theory because the estimated relationship
truly differs from the prediction, or because weak measurement, improper
solutions, poor fit, or large uncertainty limit what the structural result can
say. The measurement context stays beside the theory evidence so the two are not
confused.

# Calibration and validation

An internal split can be created explicitly:

```{r split}
s <- nomo_split(nomo_demo_network, validation_prop = .40, seed = 2026)

net_rep <- nomo_network(
  model,
  data = s,
  hypotheses = h
)

replication <- nomo_table(net_rep, "replication")
replication[, c(
  "id", "primary_estimate", "validation_estimate", "replication_status"
)]
```

```{r replication-values, include=FALSE}
rs <- function(id) replication$replication_status[replication$id == id]
# A stored status in code font, as the table shows it.
rs_code <- function(id) paste0("`", rs(id), "`")
h1_h3 <- if (identical(rs("H1"), rs("H3"))) {
  paste("both", rs_code("H1"))
} else {
  paste(rs_code("H1"), "and", rs_code("H3"))
}

h4_reading <- switch(
  rs("H4"),
  sign_change_within_uncertainty = paste(
    "neither sample's confidence interval excludes zero, so the sign change is",
    "attributed to sampling uncertainty rather than to a substantive discrepancy"
  ),
  direction_not_replicated = paste(
    "only one sample's confidence interval excludes zero, so the direction was",
    "not replicated, but no reversal is established"
  ),
  sign_reversal = paste(
    "both confidence intervals exclude zero on opposite sides, which is what a",
    "reversal requires"
  ),
  "the status is read from both confidence intervals, not the point estimates alone"
)
```

The calibration and validation subsets receive the **same prespecified fitted
model**; `nomologR` does not respecify the validation model to rescue a primary
result. Reading the statuses:

- H1 and H3 are `r h1_h3`.
- H2 is `r rs_code("H2")`: with only `r nrow(s$validation)` validation cases, the
  equivalence interval is too wide to confirm negligibility. Splitting a sample
  buys independence at the cost of precision.
- H4 is `r rs_code("H4")`. The population path is zero, and the two point
  estimates happen to differ in sign. Here `r h4_reading`.

A change in sign between samples needs care. Estimates scattered around a null
relation differ in sign about half the time, so `nomologR` does not call a sign
change a **reversal** from the point estimates. It reads the 95 percent
confidence intervals:

| Status | Intervals | Meaning |
|---|---|---|
| `sign_reversal` | Both exclude zero, on opposite sides | Each sample on its own supports a different direction |
| `direction_not_replicated` | Exactly one excludes zero | The other sample does not support that direction, nor establish the opposite |
| `sign_change_within_uncertainty` | Neither excludes zero | Neither sample distinguishes the relation from zero |

None of these turns a non-significant path into evidence of *no* relation. That
claim needs a `negligible(within = ...)` prediction with an equivalence region,
the kind of prediction H2 makes (Lakens et al., 2018).

External validation data can instead be supplied with `validation_data =`, which
is generally stronger evidence than an internal split.

# Figures

```{r effects-plot}
plot(net, type = "effects")
```

```{r replication-plot}
plot(net_rep, type = "replication")
```

These are evidence summaries, not automated theory verdicts.

# Interpretation rule

`nomologR` keeps **measurement evidence, theory concordance, uncertainty,
replication, and researcher provenance visible at the same time**. A supported
prediction strengthens the validity argument for a particular interpretation of
scores; an unsupported one is information about the theory, the measure, or
both.

# Research basis

The nomological network as the core of construct validity comes from Cronbach
and Meehl (1955), and convergent and discriminant correlation patterns from
Campbell and Fiske (1959). Contemporary practice separates the measurement model
from structural relations (Anderson & Gerbing, 1988), treats validity as an
argument about score interpretation (Messick, 1995), evaluates negligible
predictions with equivalence procedures (Schuirmann, 1987; Lakens et al.,
2018), and distinguishes prespecified from post hoc predictions (Nosek et al.,
2018). Single-indicator corrections follow Bollen (1989) and Savalei (2019).
Their standard errors follow Oberski and Satorra (2013). Full references are in
`?nomo_network`, `?nomo_single_indicator`, and the
[research basis](research-basis.html) article.
