---
title: "Archiving a nomologR workflow with nomo_report()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Archiving a nomologR workflow with nomo_report()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(nomologR)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)

# Rendering an HTML report requires pandoc. Chunks that render reports are
# skipped, with this explanation, when pandoc is unavailable.
pandoc_ok <- rmarkdown::pandoc_available()

# nomo_report() renders its own R Markdown template. Calling it from inside
# another knitr document (such as this vignette) nests two renders, so knitr
# must allow the template's chunk labels to repeat this document's labels.
# Removing the need for this setting is tracked in the issue linked below.
options(knitr.duplicate.label = "allow")
report_dir <- tempfile("nomologR-reports-")
dir.create(report_dir)
```

`nomo_report()` is the final presentation and provenance layer in the
`nomologR` workflow.

It does **not** rerun the analyses. It renders the evidence and researcher
decisions already stored in a `nomo_run` object.

## A workflow to archive

This example uses `nomo_demo_network` and supplies every consequential decision
up front, as a researcher following a preregistered plan might:

```{r run}
scales <- list(
  Agency = c("ag1", "ag2", "ag3", "ag4"),
  Persistence = c("pe1", "pe2", "pe3", "pe4")
)

h <- nomo_hypotheses(
  "Agency -> Persistence" = positive(min = .20),
  "Agency -> Performance" = positive()
)

run <- nomo_run(
  data = nomo_demo_network,
  scales = scales,
  mode = "research",
  decisions = list(
    factor_count = list(
      value = c(Agency = 1, Persistence = 1),
      rationale = "Each scale was written to measure one construct."
    ),
    cfa_model = list(
      value = nomo_model(scales),
      rationale = "Prespecified two-construct measurement model."
    ),
    measurement_model = list(
      value = "proceed",
      rationale = "Measurement evidence reviewed before planned downstream tests."
    )
  ),
  settings = list(
    factors = list(seed = 2026),
    invariance = list(group = "group", levels = c("configural", "metric", "scalar")),
    network = list(hypotheses = h)
  )
)

nomo_table(run, "stages")
nomo_table(run, "decisions")
```

## Render a report

```{r render, eval = pandoc_ok}
report_file <- nomo_report(
  run,
  file = file.path(report_dir, "construct-validation-report.html"),
  title = "Construct validation audit: Agency and Persistence"
)

file.exists(report_file)
```

The HTML file is self-contained, so it can be archived, attached to a project
record or thesis appendix, or shared with a collaborator without a separate
figures directory or a network dependency. The `title` argument becomes the
document title.

Calling `nomo_report()` from the R console or a script needs no extra setup.
Calling it from a chunk inside your own R Markdown or Quarto document (for
example, a thesis chapter) currently requires
`options(knitr.duplicate.label = "allow")` first, as this article does. Removing
that requirement is tracked in
[issue #40](https://github.com/JUhalt/nomologR/issues/40).

## What the report contains

The report mirrors the full guided workflow:

1. researcher inputs and data/sample roles;
2. item audit;
3. factor-retention evidence;
4. EFA;
5. CFA;
6. reliability;
7. convergent/discriminant evidence;
8. invariance, if requested;
9. nomological-network evidence, if requested;
10. workflow decisions and outstanding decisions;
11. deviations and post-hoc decisions;
12. methods and citations: the methods this workflow actually used, each with
    its lineage and role, a reference list for those methods, and software
    citations;
13. reproducibility information and session details.

The methods table is built from `nomo_methods(run)`, which returns only the
methods the run used; a method the package offers but the workflow did not use
is not cited. The [research basis](research-basis.html) article explains the
lineage labels.

A final evidence-trace appendix keeps each autogenerated recommendation on the
same row as the component, metric, reference, severity, and observation that
produced it.

## Reports from incomplete workflows

A report can also be rendered from a paused or blocked workflow:

```{r paused}
paused_run <- nomo_run(
  data = nomo_demo_network,
  scales = scales,
  settings = list(factors = list(seed = 2026))
)

nomo_table(paused_run, "stages")
```

```{r paused-report, eval = pandoc_ok}
paused_file <- nomo_report(
  paused_run,
  file = file.path(report_dir, "paused-workflow-audit.html"),
  include_plots = FALSE
)

file.exists(paused_file)
```

Missing stages are labeled as incomplete or not requested rather than silently
disappearing. This is useful when a researcher wants to archive why an analysis
stopped or which consequential decision remained unresolved.

## Plots and session information

The default includes a compact, nonredundant set of plots and full
`sessionInfo()` output. Both can be turned off, as in the paused report above:

```r
nomo_report(
  run,
  file = "compact-report.html",
  include_plots = FALSE,
  include_session = FALSE
)
```

## Overwriting is explicit

Existing reports are protected. Rendering to a file that already exists stops
unless `overwrite = TRUE` is supplied:

```{r overwrite, eval = pandoc_ok, error = TRUE}
nomo_report(run, file = report_file)
```

As elsewhere in `nomologR`, consequential or destructive behavior is not
silently assumed.

```{r cleanup, include=FALSE}
unlink(report_dir, recursive = TRUE)
```
