---
title: "Archiving a nomologR workflow with nomo_report()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Archiving a nomologR workflow with nomo_report()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(nomologR)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)

# Rendering an HTML report requires pandoc. Chunks that render reports are
# skipped, with this explanation, when pandoc is unavailable.
pandoc_ok <- rmarkdown::pandoc_available()

report_dir <- tempfile("nomologR-reports-")
dir.create(report_dir)
```

`nomo_report()` is the final presentation and provenance layer in the
`nomologR` workflow.

It does **not** rerun the analyses. It renders the evidence and researcher
decisions already stored in a `nomo_run` object.

## A workflow to archive

This example uses `nomo_demo_network` and supplies every consequential decision
up front, as a researcher following a preregistered plan might:

```{r run}
scales <- list(
  Agency = c("ag1", "ag2", "ag3", "ag4"),
  Persistence = c("pe1", "pe2", "pe3", "pe4")
)

h <- nomo_hypotheses(
  "Agency -> Persistence" = positive(min = .20),
  "Agency -> Performance" = positive()
)

run <- nomo_run(
  data = nomo_demo_network,
  scales = scales,
  mode = "research",
  decisions = list(
    factor_count = list(
      value = c(Agency = 1, Persistence = 1),
      rationale = "Each scale was written to measure one construct."
    ),
    cfa_model = list(
      value = nomo_model(scales),
      rationale = "Prespecified two-construct measurement model."
    ),
    measurement_model = list(
      value = "proceed",
      rationale = "Measurement evidence reviewed before planned downstream tests."
    )
  ),
  settings = list(
    factors = list(seed = 2026),
    invariance = list(group = "group", levels = c("configural", "metric", "scalar")),
    network = list(hypotheses = h)
  )
)

nomo_table(run, "stages")
nomo_table(run, "decisions")
```

## Render a report

```{r render, eval = pandoc_ok}
report_file <- nomo_report(
  run,
  file = file.path(report_dir, "construct-validation-report.html"),
  title = "Construct validation audit: Agency and Persistence"
)

file.exists(report_file)
```

The HTML file is self-contained, so it can be archived, attached to a project
record or thesis appendix, or shared with a collaborator without a separate
figures directory or a network dependency. The `title` argument becomes the
document title.

For a thesis committee or a collaborator who works in Word, give the file a
`.docx` extension instead:

```r
nomo_report(run, file = "construct-validation-report.docx")
```

The Word document carries the same tables, figures, and interpretation contract
as the HTML report. Sections that collapse in HTML, such as the full evidence
trace, are shown expanded, and it uses Word's default styles.

`nomo_report()` works the same from the console, a script, or a chunk inside
your own R Markdown or Quarto document, such as a thesis chapter. This article
renders reports from its own chunks and needs no special setup.

Rendering a report from inside another document nests two renders that share
knitr's state, so `nomo_report()` isolates them: the report is rendered with
knitr's default chunk options, which its template then sets for itself, and
your document's options are restored afterwards. Your chunk options therefore
do not change the report, and rendering the report does not change your
document.

## Tables for a manuscript

A report archives a workflow. A thesis or manuscript needs tables.
`nomo_apa_table()` formats the evidence in a result as a table in APA 7 style —
a bold number, an italic title, no vertical rules, and notes below — which knits
directly into an R Markdown or Quarto document:

```{r apa-loadings}
nomo_apa_table(run$results$cfa, "loadings", number = 1)
```

```{r apa-fit}
nomo_apa_table(run$results$cfa, "fit", number = 2)
```

```{r apa-reliability}
nomo_apa_table(run$results$reliability, number = 3)
```

```{r apa-invariance}
nomo_apa_table(run$results$invariance, number = 4)
```

```{r apa-hypotheses}
nomo_apa_table(run$results$network, number = 5)
```

Notice the leading zeros. APA 7 drops the zero before a decimal point only for a
statistic that *cannot* exceed 1, so the rule follows the statistic rather than
the value it happens to take. Reliability coefficients, correlations, CFI, and
*p* values lose it; TLI, RMSEA, SRMR, and standardized loadings keep it, because
each can exceed 1 — TLI is not bounded above, and a standardized loading does in
an improper solution. Many published tables print standardized loadings without
the zero; these tables follow the rule as written.

The notes keep the package's reference-value language. Fit indices are reported
as evidence rather than against fixed cutoffs, and the hypotheses table
describes how each estimate relates to its prediction without calling any of
them a pass or a fail. A hypothesis specified after the data were seen is marked
with a note saying it is exploratory.

## What the report contains

The report mirrors the full guided workflow:

1. researcher inputs and data/sample roles;
2. item audit;
3. factor-retention evidence;
4. EFA;
5. CFA;
6. reliability;
7. convergent/discriminant evidence;
8. invariance, if requested;
9. nomological-network evidence, if requested;
10. workflow decisions and outstanding decisions;
11. deviations and post-hoc decisions;
12. methods and citations: the methods this workflow actually used, each with
    its lineage and role, a reference list for those methods, and software
    citations;
13. reproducibility information and session details.

The methods table is built from `nomo_methods(run)`, which returns only the
methods the run used; a method the package offers but the workflow did not use
is not cited. The [research basis](research-basis.html) article explains the
lineage labels.

A final evidence-trace appendix keeps each autogenerated recommendation on the
same row as the component, metric, reference, severity, and observation that
produced it.

## Reports from incomplete workflows

A report can also be rendered from a paused or blocked workflow:

```{r paused}
paused_run <- nomo_run(
  data = nomo_demo_network,
  scales = scales,
  settings = list(factors = list(seed = 2026))
)

nomo_table(paused_run, "stages")
```

```{r paused-report, eval = pandoc_ok}
paused_file <- nomo_report(
  paused_run,
  file = file.path(report_dir, "paused-workflow-audit.html"),
  include_plots = FALSE
)

file.exists(paused_file)
```

Missing stages are labeled as incomplete or not requested rather than silently
disappearing. This is useful when a researcher wants to archive why an analysis
stopped or which consequential decision remained unresolved.

## Plots and session information

The default includes a compact, nonredundant set of plots and full
`sessionInfo()` output. Both can be turned off, as in the paused report above:

```r
nomo_report(
  run,
  file = "compact-report.html",
  include_plots = FALSE,
  include_session = FALSE
)
```

## Overwriting is explicit

Existing reports are protected. Rendering to a file that already exists stops
unless `overwrite = TRUE` is supplied:

```{r overwrite, eval = pandoc_ok, error = TRUE}
nomo_report(run, file = report_file)
```

As elsewhere in `nomologR`, consequential or destructive behavior is not
silently assumed.

```{r cleanup, include=FALSE}
unlink(report_dir, recursive = TRUE)
```
