---
title: "Scoring a measurement model with nomo_scores()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scoring a measurement model with nomo_scores()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(nomologR)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

After a measurement model is established, researchers usually need numbers to
carry into later analyses. `nomo_scores()` produces them, and reports what they
are and are not.

Scoring is a modeling decision. `nomo_scores()` documents the decision; it does
not make it.

## A model to score

```{r fit}
items <- paste0("x", 1:8)

model <- "
  Agency      =~ x1 + x2 + x3 + x4
  Persistence =~ x5 + x6 + x7 + x8
"

set.seed(33)
loadings <- cbind(
  Agency      = c(.80, .75, .70, .65, 0, 0, 0, 0),
  Persistence = c(0, 0, 0, 0, .60, .55, .50, .45)
)
phi <- matrix(c(1, .5, .5, 1), 2)
sigma <- loadings %*% phi %*% t(loadings)
diag(sigma) <- 1
dimnames(sigma) <- list(items, items)

dat <- as.data.frame(matrix(rnorm(600 * 8), 600, 8) %*% chol(sigma))
names(dat) <- items

fit <- lavaan::cfa(model, data = dat, std.lv = TRUE)
```

## Adding items is a model, not arithmetic

The most common way to score a scale is to add its items. McNeish and Wolf
(2020) show that this is not a model-free calculation: adding items assumes a
*parallel* model, in which the unstandardized loadings and the residual
variances are equal across items. That assumption needs the same justification
as any other measurement model.

`nomo_scores()` fits that constrained model and compares it with the model you
supplied:

```{r sum}
summed <- nomo_scores(fit, method = "sum")
summed
```

The loadings here were generated to differ, so the constraints are rejected.
That does not forbid a sum score. It establishes that the items are not
interchangeable in the way adding them assumes, so the choice needs a reason
beyond convenience, and the diagnostics say what it costs.

## What a score is, and is not

Grice (2001) evaluates scores on three criteria, all reported here and all
computed from the fitted model rather than from the scores themselves:

```{r diagnostics}
nomo_table(summed, "diagnostics")
```

`validity` is the correlation between a score and the factor it estimates.
Gorsuch (1983) recommended at least .80, and above .90 where scores are used as
substitutes for the factors themselves. For `method = "regression"` this
quantity is the factor determinacy coefficient that
[`nomo_hierarchical()`](hierarchical-models.html) reports, because that method
maximizes it.

`univocality` is a score's correlation with the factors it does *not*
represent. A score that is not univocal cannot be treated as though it measured
its own factor alone.

`correlational_accuracy` is the one to read before using scores in a later
analysis. It is the difference between the correlations among the scores and
the correlations among the factors they stand in for.

## Scoring changes the answer

Compare what each method implies about the relationship between the two
constructs:

```{r comparison}
comparison <- lapply(c("sum", "regression", "bartlett"), function(m) {
  s <- nomo_scores(fit, method = m)
  data.frame(
    method = m,
    factor_correlation = s$factor_correlations[1, 2],
    score_correlation = s$score_correlations[1, 2],
    discrepancy = s$diagnostics$correlational_accuracy[[1]]
  )
})

do.call(rbind, comparison)
```

The factor correlation is the same in every row, because it is a property of
the model. The score correlation is not. Unit-weighted and Bartlett scores
understate the relationship here, and regression scores overstate it.

The direction is not a constant that could be corrected for: it depends on the
scoring method and on the model. Grice reports the discrepancy running one way
for orthogonal factors and the other for oblique ones. So `nomologR` reports
it and leaves it visible rather than adjusting for it.

The practical consequence is worth stating plainly. A relationship estimated
from scores carries this discrepancy as bias. Where the question can be asked
of the latent variables instead, asking it of scores replaces an unbiased
answer with a biased one.

## Choosing a method

```{r methods}
nomo_table(nomo_scores(fit, method = "regression"), "diagnostics")
```

With one factor, regression and Bartlett scores differ only by a scaling
constant and rank people identically, so the choice between them matters only
when there is more than one factor.

Reliability coefficients follow the weighting. Coefficient alpha is the
reliability of a unit-weighted scale; coefficient H, reported by
[`nomo_hierarchical()`](hierarchical-models.html), belongs to optimally
weighted scores. Reporting one for the other describes a scale that was not
used.

## What is never done

The data you supply is never modified, and scores are returned only for the
cases the model actually used, so a case dropped from the fit does not acquire
a score.

No threshold here is applied as a rule. Gorsuch's recommendations appear where
a value falls below them, labeled as his, in the same way fixed fit-index
cutoffs are treated elsewhere in the package.

## References

Gorsuch, R. L. (1983). *Factor analysis* (2nd ed.). Lawrence Erlbaum.

Grice, J. W. (2001). Computing and evaluating factor scores. *Psychological
Methods, 6*(4), 430–450. <https://doi.org/10.1037/1082-989X.6.4.430>

McNeish, D., & Wolf, M. G. (2020). Thinking twice about sum scores. *Behavior
Research Methods, 52*(6), 2287–2305.
<https://doi.org/10.3758/s13428-020-01398-0>
