---
title: "Teaching with nomologR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Teaching with nomologR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Real data never tell a class whether an analysis found the truth. The teaching
datasets in `nomologR` do: each is simulated from a population model that is
written down in its help page, with specific problems built in. A student can
run an analysis and then check it against the answer.

This article collects exercises built on those answers. Each gives a question,
the code to run, and what the data were built to show. The answers come from
the population models, not from any one run, so they hold however the
estimates vary. The code is for students to run; it is not run here.

## Using the package in a course

* **Teaching and research modes.** `nomo_run(mode = "teaching")`, the default,
  explains each decision it asks for: the reason, the options, and the
  consequence of each. `mode = "research"` asks for the same decisions without
  the explanations. The analyses are identical in both modes.
* **Decisions as assignments.** A guided run pauses at every consequential
  choice, such as the number of factors or the measurement model, and records
  the rationale a student gives. The decision log is a record of the student's
  reasoning, which is often what an instructor wants to grade.
* **Reports as submissions.** `nomo_report()` archives a run's evidence and
  decisions in one HTML or Word document.
* **Readings.** `nomo_methods()` lists every method the package uses, with
  its references and DOIs. `nomo_methods(lineage = "historical")` lists the
  older methods students will meet in published work, and what now does their
  work. The research-basis article, `vignette("research-basis")`, tells the
  same history as a timeline.

## 1. Find the weak item and the cross-loading item

`nomo_demo_continuous` has two correlated factors with five items each. Which
items are problems, and what kind?

```{r}
library(nomologR)
fac <- nomo_factors(nomo_demo_continuous, seed = 2026)
efa <- nomo_efa(nomo_demo_continuous, factors = fac)
summary(efa)
```

<details><summary>What the data were built to show</summary>

`b5` is a weak indicator, with a population loading of .30, so it has a low
loading and little common variance. `a5` cross-loads, at .45 on A and .35 on B,
so its secondary loading is flagged. Neither flag is an instruction to delete
the item. Ask what each item's content covers, and whether the scale needs it.

</details>

## 2. How many factors?

Do the factor-retention criteria agree, and does agreement prove the count?

```{r}
fac <- nomo_factors(nomo_demo_continuous, criterion_set = "all", seed = 2026)
summary(fac)
fac <- nomo_factors(nomo_demo_continuous, criterion_set = "all",
                    missing = "complete", seed = 2026)
summary(fac)
```

<details><summary>What the data were built to show</summary>

The population has two factors. With the default pairwise handling of the
missing values, some criteria are skipped, and the summary says why: they need
one sample size for the whole correlation matrix. With complete cases, every
criterion runs, and all of them point to two.

Agreement among criteria is strong evidence for *investigating* a two-factor
solution. It is not proof, because every criterion reads the same correlation
matrix. The eigenvalue-greater-than-one rule is shown as historical context;
see `nomo_methods(lineage = "historical")` for why it is not used as evidence.

</details>

## 3. Coefficient alpha or omega?

```{r}
cfa <- nomo_cfa(nomo_model(list(A = paste0("a", 1:5), B = paste0("b", 1:5))),
                data = nomo_demo_continuous)
summary(nomo_reliability(cfa))
```

<details><summary>What the data were built to show</summary>

Alpha assumes that every item loads equally. These items don't: `b5`'s loading
is much weaker than the other B items'. Omega uses the estimated loadings, so it
does not rely on that assumption. Here the two coefficients differ only a
little, which is itself worth discussing: the assumption is violated, but the
consequence depends on how unequal the loadings are. B is less reliable than A,
largely because of `b5`.

</details>

## 4. Ordered categories

`nomo_demo_ordinal` has the same latent responses as `nomo_demo_continuous`,
cut into five ordered categories. What changes?

```{r}
fac <- nomo_factors(nomo_demo_ordinal, seed = 2026)
fac$correlation_method
cfa <- nomo_cfa(nomo_model(list(A = paste0("a", 1:5), B = paste0("b", 1:5))),
                data = nomo_demo_ordinal, ordered = names(nomo_demo_ordinal))
cfa
```

<details><summary>What the data were built to show</summary>

The structure is the same, but ordered items call for polychoric correlations
and an estimator for categorical data (WLSMV). The package chooses polychoric
correlations for ordered factors and never converts category labels to numbers
silently. Compare the loadings with those from the continuous version.

</details>

## 5. A correlation is not a path

In `nomo_demo_network`, Persistence and Performance are correlated. Does
Persistence predict Performance?

```{r}
scales <- list(
  Agency = paste0("ag", 1:4),
  Persistence = paste0("pe", 1:4),
  SocialDesirability = paste0("sd", 1:3)
)
h <- nomo_hypotheses(
  "Agency -> Performance" = positive(),
  "Persistence -> Performance" = negligible(within = c(-.10, .10)),
  "Agency -> Persistence" = positive(min = .20)
)
net <- nomo_network(nomo_model(scales), data = nomo_demo_network, hypotheses = h)
summary(net)
```

<details><summary>What the data were built to show</summary>

Performance depends on Agency (.40) and not on Persistence (0). Persistence
depends on Agency (.45), so Persistence and Performance are still correlated,
about .18 in the population, through their shared cause. The structural path
answers a different question from the correlation.

The Persistence path is near zero and not significant, yet the `negligible()`
prediction is not supported: its interval still reaches beyond the region of
-.10 to .10 stated in advance. A path that is not significant is not evidence of no
effect. Establishing that an effect is negligible takes an interval inside the
region, which usually needs more precision than detecting an effect does. Ask
students what sample size would be needed.

</details>

## 6. Find the item that is not comparable across groups

The data were collected online and on paper. Can scores be compared across the
two groups?

```{r}
inv <- nomo_invariance("Agency =~ ag1 + ag2 + ag3 + ag4",
                       data = nomo_demo_network, group = "group",
                       levels = c("configural", "metric", "scalar"))
summary(inv)
```

<details><summary>What the data were built to show</summary>

Loadings are equal across groups, so metric invariance holds. The intercept of
`ag3` is .50 higher on paper, so equal intercepts cost fit, and the largest
score diagnostic is `ag3`'s intercept. The latent Agency mean is also .25 higher
on paper. That is a real difference, which an unequal intercept would distort.
Whether to release the intercept is the researcher's decision, with a
rationale (`nomo_partial()`).

</details>

## 7. When content review and the data disagree

`nomo_demo_walkthrough` is a twelve-item scale an expert panel reviewed in
`contentvalidR`. The review carried ten items. Screen them.

```{r}
handoff <- readRDS(system.file("extdata", "content-handoff-walkthrough.rds",
                               package = "nomologR"))
run <- nomo_run(nomo_demo_walkthrough, scales = handoff)
run
summary(run$results$screen$EF)
```

<details><summary>What the data were built to show</summary>

As the data arrive, every item is flagged. `EF2` and `TF2` are reverse-worded
and not yet recoded, so they run against their scales and pull every other
item's correlation down. The audit says so. Recode them (6 minus the response)
and screen again. `EF4` is the one Effort Regulation item still flagged: the
panel placed it, and it carries almost no common variance. `EF3` passes the
screen, but its answers pile up at the ceiling, and the measurement model later
gives it a weak loading. It is also the only item about finishing required
work. `vignette("content-review")` works through the full example, including
`TF4`, which loads on both facets, and `TF6`, which differs by cohort.

</details>
